---
title: "Climate Departure Analysis for a Watershed"
output: bookdown::html_document2
vignette: >
  %\VignetteIndexEntry{Climate Departure Analysis for a Watershed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r libraries, include = FALSE}
library(terra)
library(ggplot2)
library(tidyterra)
```

The cd package computes climate departure statistics from ERA5-Land
reanalysis data for any area of interest. This vignette demonstrates the
full analysis pipeline: connecting to the cloud-hosted data catalog, extracting
zonal means for a watershed, computing anomalies against a pre-warming
baseline, and running trend analysis.

## Area of Interest

We use the Kootenay Lake (KOTL) watershed group from the BC Freshwater
Atlas, shipped with the package. Any `sf` polygon works as an AOI.

```{r load-aoi}
library(cd)
library(sf)

aoi <- st_read(
  system.file("extdata", "example_aoi_kotl.gpkg", package = "cd"),
  quiet = TRUE
)
```

```{r load-context, include = FALSE}
ctx <- system.file("extdata", "context_kotl.gpkg", package = "cd")
towns <- st_read(ctx, layer = "towns", quiet = TRUE)
lakes <- st_read(ctx, layer = "lakes", quiet = TRUE)
rivers <- st_read(ctx, layer = "rivers", quiet = TRUE)
streams <- st_read(ctx, layer = "streams", quiet = TRUE)
highways <- st_read(ctx, layer = "highways", quiet = TRUE)

# Clip context layers to the watershed group (AOI) so maps stay focused
# on the basin and do not show features outside the boundary. Mark
# attributes as spatially constant to silence the (irrelevant) attribute-
# constancy notice without hiding genuine geometry warnings.
for (lyr in list("towns", "lakes", "rivers", "streams", "highways")) {
  x <- get(lyr)
  sf::st_agr(x) <- "constant"
  assign(lyr, sf::st_intersection(x, aoi))
}
```

```{r map-location, fig.cap="Kootenay Lake (KOTL) watershed group in southeastern British Columbia showing major waterbodies, rivers, and highways.", fig.height=7}
aoi_bb <- sf::st_bbox(aoi)

ggplot() +
  geom_sf(data = aoi, fill = "#f0f0f0", color = "#2166ac", linewidth = 0.8) +
  geom_sf(data = lakes, fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = rivers, fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = streams, color = "#74a9cf", linewidth = 0.4, alpha = 0.6) +
  geom_sf(data = highways, color = "#333333", linewidth = 0.8) +
  geom_sf(data = towns, color = "black", size = 2.5) +
  ggrepel::geom_label_repel(
    data = towns,
    aes(label = name, geometry = geom),
    stat = "sf_coordinates",
    size = 3.5, fill = "white", alpha = 0.85,
    label.padding = unit(0.2, "lines"),
    min.segment.length = 0
  ) +
  coord_sf(
    xlim = c(aoi_bb["xmin"], aoi_bb["xmax"]),
    ylim = c(aoi_bb["ymin"], aoi_bb["ymax"])
  ) +
  labs(title = "Kootenay Lake Watershed Group (KOTL)") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank())
```

## Connect to the Data Catalog

The cd package serves ERA5-Land climate data as Cloud-Optimized GeoTIFFs
on S3, indexed by a STAC catalog. `cd_catalog()` reads the catalog and
returns a tidy tibble of available variables and periods.

```{r catalog}
catalog <- cd_catalog()
knitr::kable(catalog, caption = "Available climate variables and periods in the STAC catalog.")
```

## Extract Climate Time Series

`cd_extract()` crops each COG to the AOI and computes the spatial mean
per year. This runs directly against the cloud-hosted data — no local
download needed.

```{r extract}
ts <- cd_extract(catalog, aoi)
knitr::kable(head(ts, 10), caption = "First 10 rows of the extracted climate time series.")
```

We now have annual and seasonal values for five climate variables across
76 years (1950--2025) for the Kootenay Lake watershed.

## Choosing a Baseline

The choice of reference period shapes the story. The WMO standard
1981--2010 baseline is widely used for comparability, but it includes
decades of warming. A pre-warming baseline (1951--1980) reveals the
full magnitude of departure — what Pauly (1995) called avoiding the
"shifting baseline syndrome."

```{r baseline}
# Pre-warming baseline
bl_early <- cd_baseline(ts, baseline_years = 1951:1980)

# WMO standard
bl_wmo <- cd_baseline(ts, baseline_years = 1981:2010)

knitr::kable(bl_early, caption = "Pre-warming baseline means (1951-1980) by variable and period.", digits = 2)
```

## Anomalies

`cd_anomaly()` computes departures from the baseline. Temperature, VPD,
and RH use absolute deviations. Precipitation and soil moisture use
percent of normal.

```{r anomaly}
ano <- cd_anomaly(ts, bl_early)
knitr::kable(head(ano, 10), caption = "Anomalies relative to the 1951-1980 baseline.", digits = 3)
```

## Temperature Departure

```{r plot-tmean, fig.cap="Annual mean temperature anomaly for the Kootenay Lake watershed relative to 1951-1980 baseline. Red bars indicate warmer-than-baseline years."}
trn <- cd_trend(ano, trend_start = c(1951, 1981))
cd_plot_timeseries(
  ano,
  variable = "tmean",
  period = "annual",
  trend = trn,
  title = "Mean Temperature Anomaly — Kootenay Lake Watershed"
)
```

```{r summary-tmean}
knitr::kable(cd_summary(trn[trn$variable == "tmean", ]),
  caption = "Temperature trend statistics for the Kootenay Lake watershed.")
```

The Kootenay Lake watershed has warmed substantially since the mid-20th
century. The annual mean temperature trend since 1951 gives a Total
Change that quantifies this cumulative shift.

## Comparing Time Windows

`cd_compare()` directly answers: "How different is the recent climate
from the historical climate?"

```{r compare}
cmp <- cd_compare(ts,
  window_a = 2015:2025,
  window_b = 1951:1980,
  method = "mean_diff"
)
knitr::kable(cmp, caption = "Comparison of recent (2015-2025) vs pre-warming (1951-1980) means.", digits = 2)
```

```{r plot-compare, fig.cap="Comparison of recent (2015-2025) vs pre-warming (1951-1980) climate for the Kootenay Lake watershed."}
cd_plot_comparison(
  cmp,
  labels = c(a = "2015-2025", b = "1951-1980"),
  title = "Climate Shift — Kootenay Lake Watershed"
)
```

## Spatial Patterns of Departure

The zonal mean tells us the watershed average, but departure varies
across the landscape. Valley bottoms, high elevation areas, and rain
shadows respond differently. We can map this by computing the difference
between recent and historical period means directly from the rasters.

```{r spatial-tmean, fig.cap="Spatial pattern of annual mean temperature departure across the Kootenay Lake watershed. Difference between 2015-2025 mean and 1951-1980 mean (degrees C).", fig.height=6}
# Read the annual tmean COG and crop to AOI
tmean_row <- catalog[catalog$variable == "tmean" & catalog$period == "annual", ]
r_tmean <- cd_crop(tmean_row$href, aoi)

# Compute period means from the multi-year raster bands
years <- as.integer(names(r_tmean))
recent_idx <- which(years >= 2015 & years <= 2025)
historical_idx <- which(years >= 1951 & years <= 1980)

recent_mean <- mean(r_tmean[[recent_idx]])
historical_mean <- mean(r_tmean[[historical_idx]])
departure <- recent_mean - historical_mean
# Mask to the watershed boundary so cells outside the AOI go transparent
departure <- terra::mask(departure, aoi)
names(departure) <- "Temperature departure"

ggplot() +
  geom_spatraster(data = departure) +
  geom_sf(data = aoi, fill = NA, color = "black", linewidth = 0.6) +
  geom_sf(data = lakes, fill = NA, color = "grey40", linewidth = 0.2) +
  geom_sf(data = highways, color = "#333333", linewidth = 0.5) +
  geom_sf(data = towns, color = "black", size = 2) +
  ggrepel::geom_label_repel(
    data = towns, aes(label = name, geometry = geom),
    stat = "sf_coordinates", size = 3, fill = "white", alpha = 0.8
  ) +
  scale_fill_distiller(
    palette = "RdBu", direction = -1,
    name = expression(Delta * degree * C)
  ) +
  coord_sf(
    xlim = c(aoi_bb["xmin"], aoi_bb["xmax"]),
    ylim = c(aoi_bb["ymin"], aoi_bb["ymax"])
  ) +
  labs(title = "Temperature Departure (2015-2025 vs 1951-1980)") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank())
```

```{r spatial-sm, fig.cap="Spatial pattern of summer soil moisture departure. Difference between 2015-2025 mean and 1951-1980 mean (m3/m3). Negative values (brown) indicate drying.", fig.height=6}
sm_row <- catalog[catalog$variable == "soil_moisture" & catalog$period == "summer", ]
r_sm <- cd_crop(sm_row$href, aoi)

years_sm <- as.integer(names(r_sm))
recent_sm <- mean(r_sm[[which(years_sm >= 2015 & years_sm <= 2025)]])
historical_sm <- mean(r_sm[[which(years_sm >= 1951 & years_sm <= 1980)]])
departure_sm <- recent_sm - historical_sm
# Mask to the watershed boundary so cells outside the AOI go transparent
departure_sm <- terra::mask(departure_sm, aoi)
names(departure_sm) <- "Soil moisture departure"

ggplot() +
  geom_spatraster(data = departure_sm) +
  geom_sf(data = aoi, fill = NA, color = "black", linewidth = 0.6) +
  geom_sf(data = lakes, fill = NA, color = "grey40", linewidth = 0.2) +
  geom_sf(data = highways, color = "#333333", linewidth = 0.5) +
  geom_sf(data = towns, color = "black", size = 2) +
  ggrepel::geom_label_repel(
    data = towns, aes(label = name, geometry = geom),
    stat = "sf_coordinates", size = 3, fill = "white", alpha = 0.8
  ) +
  scale_fill_distiller(
    palette = "BrBG", direction = 1,
    name = expression(Delta ~ m^3/m^3)
  ) +
  coord_sf(
    xlim = c(aoi_bb["xmin"], aoi_bb["xmax"]),
    ylim = c(aoi_bb["ymin"], aoi_bb["ymax"])
  ) +
  labs(title = "Summer Soil Moisture Departure (2015-2025 vs 1951-1980)") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank())
```

## Seasonal Patterns

Temperature warming is often strongest in specific seasons. Let's look
at summer and winter separately.

```{r seasonal}
summer_ano <- ano[ano$period == "summer" & ano$variable == "tmean", ]
winter_ano <- ano[ano$period == "winter" & ano$variable == "tmean", ]

trn_summer <- cd_trend(summer_ano, trend_start = 1951)
trn_winter <- cd_trend(winter_ano, trend_start = 1951)

cd_plot_timeseries(summer_ano, period = "summer", trend = trn_summer,
  title = "Summer Temperature Anomaly")
```

```{r seasonal-summary}
knitr::kable(cd_summary(rbind(trn_summer, trn_winter)),
  caption = "Seasonal temperature trends (summer vs winter).")
```

## Day-Night Asymmetry

Mean temperature smooths over a signal that matters ecologically: in most
of British Columbia's interior, **overnight lows (tmin) have warmed faster
than daytime highs (tmax)** since the mid-20th century. This day-night
asymmetry is one of the clearest fingerprints of greenhouse warming
(Karl et al. 1993; IPCC AR6) — and it is precisely the signal that
salmon, soils, and snowpack respond to.

The cd package now ships tmax and tmin alongside tmean. Pulling the same
extracted time series and plotting them separately reveals the asymmetry
that the mean alone hides.

```{r plot-tmax, fig.cap="Annual daytime maximum temperature (tmax) anomaly for the Kootenay Lake watershed relative to the 1951-1980 baseline."}
trn_tmax <- cd_trend(
  ano[ano$variable == "tmax" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "tmax", period = "annual", trend = trn_tmax,
  title = "Daytime Maximum (tmax) — Annual Anomaly")
```

```{r plot-tmin, fig.cap="Annual overnight minimum temperature (tmin) anomaly for the Kootenay Lake watershed relative to the 1951-1980 baseline. Compare the trend slope to tmax above to see the day-night asymmetry."}
trn_tmin <- cd_trend(
  ano[ano$variable == "tmin" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "tmin", period = "annual", trend = trn_tmin,
  title = "Overnight Minimum (tmin) — Annual Anomaly")
```

The Theil-Sen slopes typically show tmin warming faster than tmax for
interior BC watersheds. The implication is that the **diurnal temperature
range** (DTR = tmax − tmin) is shrinking — nights are catching up to
days. We can compute and plot DTR directly from the extracted series.

```{r plot-dtr, fig.cap="Diurnal temperature range (tmax minus tmin) annual mean for the Kootenay Lake watershed. A declining trend indicates that overnight lows are warming faster than daytime highs."}
tmax_ts <- ts[ts$variable == "tmax" & ts$period == "annual", c("year", "value")]
tmin_ts <- ts[ts$variable == "tmin" & ts$period == "annual", c("year", "value")]
dtr <- merge(tmax_ts, tmin_ts, by = "year", suffixes = c("_max", "_min"))
dtr$dtr <- dtr$value_max - dtr$value_min

ggplot(dtr, aes(x = year, y = dtr)) +
  geom_line(color = "grey50") +
  geom_point(color = "grey30", size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "#b2182b", linewidth = 0.8) +
  labs(
    title = "Diurnal Temperature Range — Kootenay Lake Watershed",
    x = NULL,
    y = expression("DTR (" * degree * "C)")
  ) +
  theme_minimal(base_size = 12)
```

A shrinking DTR has cascading consequences:

- **Evapotranspiration intensifies.** Warmer nights mean less overnight
  cooling, raising daily-mean vapour pressure deficit and pulling more
  water from soils and plants. This compounds the soil-drying signal
  seen in the next section.
- **Cold-air pooling weakens.** Valley-bottom inversions that historically
  trapped frost in the growing season are eroded, shifting frost-free
  windows and altering pest range — for example, mountain pine beetle
  overwintering survival is governed by tmin extremes.
- **Stream thermal regime shifts.** While salmon thermal stress is
  primarily an afternoon-tmax problem, summer tmin sets the *minimum*
  daily temperature that streams cool back toward overnight, raising the
  daily-mean substrate temperature that drives metabolism.

## Precipitation and Soil Moisture

While temperature shows clear departure, precipitation trends in BC
are often less pronounced. Soil moisture integrates both temperature
and precipitation signals — warmer temperatures drive more
evapotranspiration even when rainfall is stable.

```{r soil-moisture}
sm_ano <- ano[ano$variable == "soil_moisture" & ano$period == "summer", ]
trn_sm <- cd_trend(sm_ano, trend_start = 1951)

cd_plot_timeseries(sm_ano, variable = "soil_moisture", period = "summer",
  trend = trn_sm, title = "Summer Soil Moisture Anomaly")
```

```{r all-trends}
all_trends <- cd_trend(
  ano[ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
knitr::kable(cd_summary(all_trends),
  caption = "Annual trend statistics for all variables, Kootenay Lake watershed.")
```

## Interpretation

The analysis reveals the climate departure pattern common across British
Columbia's interior watersheds:

- **Temperature is rising.** The cumulative shift since the mid-20th century
  is substantial and statistically significant across all seasons.
- **Precipitation trends are weak.** Year-to-year variability is high but
  there is no strong directional change.
- **Soils are drying despite stable precipitation.** Warmer temperatures
  increase evapotranspiration, pulling moisture from soils even when the
  same amount of rain falls. For aquatic ecosystems, drier soils mean
  reduced summer baseflows during the period when flows are already lowest.

This pattern — warming without precipitation change leading to hydrological
drought — is the mechanism connecting climate departure to habitat
degradation in salmon-bearing watersheds.

## Data Source

All data are from the [ERA5-Land](https://www.ecmwf.int/en/era5-land)
reanalysis dataset produced by ECMWF for the Copernicus Climate Change
Service (Muñoz-Sabater et al. 2021). Anomalies are relative to user-defined
baseline periods. Trends are computed using the Mann-Kendall test for
significance and the Theil-Sen estimator for slope magnitude.
