---
title: "Climate Departure Analysis for a Watershed"
output: bookdown::html_document2
vignette: >
  %\VignetteIndexEntry{Climate Departure Analysis for a Watershed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

The cd package computes climate departure statistics from ERA5-Land
reanalysis data for any area of interest. This vignette demonstrates the
full analysis pipeline: connecting to the cloud-hosted data catalog, extracting
zonal means for a watershed, computing anomalies against a pre-warming
baseline, and running trend analysis.

## Area of Interest

We use the Kootenay Lake (KOTL) watershed group from the BC Freshwater
Atlas, shipped with the package. Any `sf` polygon works as an AOI.

```{r load-aoi}
library(cd)
library(sf)

aoi <- st_read(
  system.file("extdata", "example_aoi_kotl.gpkg", package = "cd"),
  quiet = TRUE
)
```

## Connect to the Data Catalog

The cd package serves ERA5-Land climate data as Cloud-Optimized GeoTIFFs
on S3, indexed by a STAC catalog. `cd_catalog()` reads the catalog and
returns a tidy tibble of available variables and periods.

```{r catalog}
catalog <- cd_catalog()
catalog
```

## Extract Climate Time Series

`cd_extract()` crops each COG to the AOI and computes the spatial mean
per year. This runs directly against the cloud-hosted data — no local
download needed.

```{r extract}
ts <- cd_extract(catalog, aoi)
head(ts, 10)
```

We now have annual and seasonal values for five climate variables across
76 years (1950--2025) for the Kootenay Lake watershed.

## Choosing a Baseline

The choice of reference period shapes the story. The WMO standard
1981--2010 baseline is widely used for comparability, but it includes
decades of warming. A pre-warming baseline (1951--1980) reveals the
full magnitude of departure — what Pauly (1995) called avoiding the
"shifting baseline syndrome."

```{r baseline}
# Pre-warming baseline
bl_early <- cd_baseline(ts, baseline_years = 1951:1980)

# WMO standard
bl_wmo <- cd_baseline(ts, baseline_years = 1981:2010)

bl_early
```

## Anomalies

`cd_anomaly()` computes departures from the baseline. Temperature, VPD,
and RH use absolute deviations. Precipitation and soil moisture use
percent of normal.

```{r anomaly}
ano <- cd_anomaly(ts, bl_early)
head(ano)
```

## Temperature Departure

```{r plot-tmean, fig.cap="Annual mean temperature anomaly for the Kootenay Lake watershed relative to 1951-1980 baseline. Red bars indicate warmer-than-baseline years."}
trn <- cd_trend(ano, trend_start = c(1951, 1981))
cd_plot_timeseries(
  ano,
  variable = "tmean",
  period = "annual",
  trend = trn,
  title = "Mean Temperature Anomaly — Kootenay Lake Watershed"
)
```

```{r summary-tmean}
cd_summary(trn[trn$variable == "tmean", ])
```

The Kootenay Lake watershed has warmed substantially since the mid-20th
century. The annual mean temperature trend since 1951 gives a Total
Change that quantifies this cumulative shift.

## Comparing Time Windows

`cd_compare()` directly answers: "How different is the recent climate
from the historical climate?"

```{r compare}
cmp <- cd_compare(ts,
  window_a = 2015:2025,
  window_b = 1951:1980,
  method = "mean_diff"
)
cmp
```

```{r plot-compare, fig.cap="Comparison of recent (2015-2025) vs pre-warming (1951-1980) climate for the Kootenay Lake watershed."}
cd_plot_comparison(
  cmp,
  labels = c(a = "2015-2025", b = "1951-1980"),
  title = "Climate Shift — Kootenay Lake Watershed"
)
```

## Seasonal Patterns

Temperature warming is often strongest in specific seasons. Let's look
at summer and winter separately.

```{r seasonal}
summer_ano <- ano[ano$period == "summer" & ano$variable == "tmean", ]
winter_ano <- ano[ano$period == "winter" & ano$variable == "tmean", ]

trn_summer <- cd_trend(summer_ano, trend_start = 1951)
trn_winter <- cd_trend(winter_ano, trend_start = 1951)

cd_plot_timeseries(summer_ano, period = "summer", trend = trn_summer,
  title = "Summer Temperature Anomaly")
```

```{r seasonal-summary}
cd_summary(rbind(trn_summer, trn_winter))
```

## Precipitation and Soil Moisture

While temperature shows clear departure, precipitation trends in BC
are often less pronounced. Soil moisture integrates both temperature
and precipitation signals — warmer temperatures drive more
evapotranspiration even when rainfall is stable.

```{r soil-moisture}
sm_ano <- ano[ano$variable == "soil_moisture" & ano$period == "summer", ]
trn_sm <- cd_trend(sm_ano, trend_start = 1951)

cd_plot_timeseries(sm_ano, variable = "soil_moisture", period = "summer",
  trend = trn_sm, title = "Summer Soil Moisture Anomaly")
```

```{r all-trends}
all_trends <- cd_trend(
  ano[ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_summary(all_trends)
```

## Interpretation

The analysis reveals the climate departure pattern common across British
Columbia's interior watersheds:

- **Temperature is rising.** The cumulative shift since the mid-20th century
  is substantial and statistically significant across all seasons.
- **Precipitation trends are weak.** Year-to-year variability is high but
  there is no strong directional change.
- **Soils are drying despite stable precipitation.** Warmer temperatures
  increase evapotranspiration, pulling moisture from soils even when the
  same amount of rain falls. For aquatic ecosystems, drier soils mean
  reduced summer baseflows during the period when flows are already lowest.

This pattern — warming without precipitation change leading to hydrological
drought — is the mechanism connecting climate departure to habitat
degradation in salmon-bearing watersheds.

## Data Source

All data are from the [ERA5-Land](https://www.ecmwf.int/en/era5-land)
reanalysis dataset produced by ECMWF for the Copernicus Climate Change
Service (Muñoz-Sabater et al. 2021). Anomalies are relative to user-defined
baseline periods. Trends are computed using the Mann-Kendall test for
significance and the Theil-Sen estimator for slope magnitude.
