---
title: "Climate Departure for the Kootenay Lake Region"
output:
  bookdown::html_document2:
    number_sections: false
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Climate Departure for the Kootenay Lake Region}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
After adding or removing pandoc citation markers in this vignette
(or in peace-fwcp.Rmd), regenerate vignettes/references.bib by
running:

  Rscript data-raw/regenerate_bib.R

Pulls source records from Zotero via Better BibTeX (rbbt -> BBT).
Requires Zotero running with the BBT plugin enabled. CI does not
run this script — the bib is committed and pkgdown reads it.
-->


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r libraries, include = FALSE}
library(terra)
library(ggplot2)
library(tidyterra)
```

The `cd` package builds on ERA5-Land hourly reanalysis (1950–present,
~9 km native grid) fetched from the DestinE Earth Data Hub, aggregated
to monthly, seasonal, and annual periods over British Columbia, and
published as Cloud-Optimized GeoTIFFs alongside a static STAC catalog
in a public S3 bucket. R consumer functions read those GeoTIFFs
directly via GDAL, crop to any area of interest, and compute baselines,
anomalies, and Mann-Kendall / Theil-Sen trend statistics.

This vignette runs the consumer pipeline on a four-watershed-group
area in the southern interior of British Columbia: Kootenay Lake
(`KOTL`), Lower Arrow Lake (`LARL`, which covers Trail / Rossland /
Red Mountain at its south end), Duncan Lake (`DUNC`, draining the
north end of Kootenay Lake including the Lardeau valley), and Slocan
River (`SLOC`, between the Selkirks and Monashees). The four
watershed groups together total ~24,000 km² and span the east-west
precipitation gradient that defines the snowpack story for the
region — Selkirk Pacific spillover west of Kootenay Lake, Purcell
rain shadow east.

## Area of Interest

The bundled area of interest is the union of the four watershed
groups (single multi-polygon, EPSG:4326). Any `sf` polygon works. The
watersheds in this region all drain to the Columbia River — Kootenay
Lake feeds the Kootenay River which joins the Columbia at Castlegar;
the Slocan and Lower Arrow drain directly into the Columbia.

We also carry British Columbia ecoregions through the analysis.
Ecoregions partition the province into broad climate-physiography
zones based on latitude, elevation, and dominant vegetation. Climate
departure can vary across them in ways the regional average hides,
and we use them later as the sub-region for the per-ecoregion
breakdown.

```{r load-aoi, echo = FALSE}
library(cd)
library(sf)

aoi <- st_read(
  system.file("extdata", "example_aoi_kootenay_lake.gpkg", package = "cd"),
  quiet = TRUE
)

area_km2 <- as.numeric(sum(st_area(st_transform(aoi, 3005)))) / 1e6
```

```{r load-context, include = FALSE}
ctx <- system.file("extdata", "context_kootenay_lake.gpkg", package = "cd")
towns       <- st_read(ctx, layer = "towns",      quiet = TRUE)
lakes       <- st_read(ctx, layer = "lakes",      quiet = TRUE)
rivers      <- st_read(ctx, layer = "rivers",     quiet = TRUE)
streams     <- st_read(ctx, layer = "streams",    quiet = TRUE)
highways    <- st_read(ctx, layer = "highways",   quiet = TRUE)
ecoregions  <- st_read(ctx, layer = "ecoregions", quiet = TRUE)
ecoregions  <- ecoregions[ecoregions$area_km2 > 50, ]
ecoregions$name_tc <- tools::toTitleCase(tolower(ecoregions$name))

wsgs <- st_read(ctx, layer = "wsgs", quiet = TRUE)
```

```{r map-location, fig.cap="Kootenay Lake Region area of interest (~24,000 km²) in southern interior British Columbia, coloured by ecoregion. Kootenay Lake dominates the basin; Lower Arrow Lake and the Slocan are visible to the west.", fig.height=7, echo = FALSE}
aoi_bb <- st_bbox(aoi)

ggplot() +
  geom_sf(data = ecoregions, aes(fill = name_tc), color = "grey45",
          linewidth = 0.3, alpha = 0.45) +
  geom_sf(data = aoi, fill = NA, color = "#2166ac", linewidth = 0.8) +
  geom_sf(data = lakes, fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = rivers, fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = streams, color = "#74a9cf", linewidth = 0.3, alpha = 0.6) +
  # Highway: grey casing under warm yellow fill (gq reg_qgis_restoration)
  geom_sf(data = highways, color = "#9c9c9c", linewidth = 0.75) +
  geom_sf(data = highways, color = "#ffe585", linewidth = 0.5) +
  geom_sf(data = towns, color = "black", size = 2.5) +
  ggrepel::geom_label_repel(
    data = towns,
    aes(label = name, geometry = geom),
    stat = "sf_coordinates",
    size = 3.2, fill = "white", alpha = 0.85,
    label.padding = unit(0.2, "lines"),
    min.segment.length = 0
  ) +
  scale_fill_brewer(palette = "Set3", name = "Ecoregion") +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_sf(
    xlim = c(aoi_bb["xmin"] - 0.4, aoi_bb["xmax"] + 0.4),
    ylim = c(aoi_bb["ymin"] - 0.5, aoi_bb["ymax"] + 0.3)
  ) +
  labs(title = "Kootenay Lake Region",
       subtitle = paste0(round(area_km2), " km² — ", nrow(ecoregions), " ecoregions")) +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.key.size = unit(0.4, "cm"))
```

## Connect to the Data Catalog

The producer pipeline in this repository fetches ERA5-Land hourly
reanalysis data, derives variables (vapour pressure deficit, relative
humidity, soil moisture), aggregates to monthly / seasonal / annual,
and writes Cloud-Optimized GeoTIFFs to S3. Alongside the GeoTIFFs we
publish a static SpatioTemporal Asset Catalog (STAC) — a small set of
JSON files that index the assets. Both the GeoTIFFs and the catalog
JSON live at
<https://stac-era5-land.s3.us-west-2.amazonaws.com/catalog.json>.

`cd_catalog()` reads that URL by default. The Cloud-Optimized GeoTIFFs
are also usable directly outside R — for example, in QGIS via the STAC
plugin, in `gdalcubes`, or with any STAC-aware client.

```{r catalog}
catalog <- cd_catalog()
kableExtra::kable_styling(
  knitr::kable(catalog, label = NA,
    caption = "Available climate variables and periods in the STAC catalog."),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "320px")
```

<br>

## Extract Climate Time Series

`cd_extract()` crops each cloud-hosted GeoTIFF to the area of interest
and computes the spatial mean per year. For an area of interest of
this size a live extraction takes a few seconds per variable. To keep
the vignette fast and reproducible we load a pre-computed result of
exactly that call.

```{r extract-recipe, eval = FALSE}
ts <- cd_extract(catalog, aoi)
head(ts, 10)
```

```{r extract-load, echo = FALSE}
# Load the pre-computed equivalent so the vignette renders without
# hitting the network. ts is the time series tibble; bl_early, ano,
# trn, cmp, and cmp_pct are the downstream products of the cd
# pipeline (computed once in data-raw/kootenay_lake_vignette_data.R).
vignette_data <- readRDS(system.file(
  "vignette-data", "kootenay_lake.rds", package = "cd"
))
ts       <- vignette_data$regional$ts
bl_early <- vignette_data$regional$bl
ano      <- vignette_data$regional$ano
trn      <- vignette_data$regional$trn
cmp      <- vignette_data$regional$cmp
cmp_pct  <- vignette_data$regional$cmp_pct

knitr::kable(head(ts, 10),
  caption = "First 10 rows of the extracted climate time series.")
```

## Trends

Anomalies are computed against a pre-warming reference period — 1951–1980,
the three decades before climate change accelerated. This is the same base
period @hansen_etal2012Perceptionclimate use to detect the emergence of
3-sigma summertime-temperature outliers globally. Saying a year is
"+1.5 °C" means it was 1.5 °C warmer than the average year between 1951
and 1980.

The trend table that follows has two rows per variable. We compute trends
from two different start years:

- **1951–present (75 years)** — the long view. Captures the full
  magnitude of warming since the pre-warming reference.
- **1981–present (45 years)** — starts at the beginning of the World
  Meteorological Organization's most recent 30-year "climate normal"
  (1981–2010). This is the reference period used in most published
  climate products, so it makes results easy to compare against
  Intergovernmental Panel on Climate Change reports and government
  climate summaries.

Comparing the two slopes is informative. If the 45-year slope is steeper
than the 75-year slope, warming has accelerated — recent decades are
heating faster than the long-term average. If the 45-year slope is
shallower, warming has slowed (though "slower" almost never means
"stopped"). When the two slopes are similar, the rate of change has been
roughly steady across the full record.

Total Change is the slope multiplied by the number of years — the
cumulative shift over the trend window.

```{r trend-recipe, eval = FALSE}
bl_early <- cd_baseline(ts, baseline_years = 1951:1980)
ano      <- cd_anomaly(ts, bl_early)
trn      <- cd_trend(ano, trend_start = c(1951, 1981))
cd_summary(trn)
```

```{r trend-table, echo = FALSE}
kableExtra::kable_styling(
  knitr::kable(cd_summary(trn), label = NA,
    caption = "Trend statistics for all variables and periods, Kootenay Lake Region."),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "320px")
```

<br>

```{r plot-tmean, fig.cap="Annual mean temperature anomaly for the Kootenay Lake Region relative to 1951-1980 baseline."}
cd_plot_timeseries(
  ano, variable = "tmean", period = "annual", trend = trn,
  title = "Annual Mean Temperature Anomaly — Kootenay Lake Region"
)
```

```{r plot-prcp, fig.cap="Annual precipitation anomaly (% of 1951-1980 baseline) for the Kootenay Lake Region."}
cd_plot_timeseries(
  ano, variable = "prcp", period = "annual", trend = trn,
  title = "Annual Precipitation Anomaly — Kootenay Lake Region"
)
```

## Daytime Highs and Overnight Lows

The cd package ships daytime maximum (tmax) and overnight minimum (tmin)
temperatures alongside the daily mean. They carry distinct information.
Overnight minimums warming faster than daytime maximums — the
"day-night asymmetry" — is one of the textbook fingerprints of
greenhouse warming. @karl_etal1993NewPerspective documented this
first at the global scale, finding overnight minimums rose at
roughly three times the rate of daytime maximums between 1951 and
1990 (0.84 vs 0.28 °C). Whether a watershed or region shows that
signal depends on local geography (valley inversions, snow cover,
slope-aspect mix).

For the Kootenay Lake Region, **overnight minimums are warming faster
than daytime maximums** — the textbook day-night asymmetry, though
the gap here is smaller than at higher-latitude sites. Daytime
maximums warmed about +0.024 °C per year since 1951 (+1.8 °C
cumulative), while overnight minimums warmed about +0.027 °C per year
(+2.0 °C cumulative). The overnight side warmed roughly 0.2 °C more
than the daytime side over the full record. The three figures below
show the tmax, tmin and diurnal-range time series that yield those
numbers.

```{r plot-tmax, fig.cap="Annual daytime maximum temperature (tmax) anomaly for the Kootenay Lake Region relative to the 1951-1980 baseline."}
trn_tmax <- cd_trend(
  ano[ano$variable == "tmax" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "tmax", period = "annual", trend = trn_tmax,
  title = "Daytime Maximum (tmax) — Annual Anomaly")
```

```{r plot-tmin, fig.cap="Annual overnight minimum temperature (tmin) anomaly for the Kootenay Lake Region relative to the 1951-1980 baseline."}
trn_tmin <- cd_trend(
  ano[ano$variable == "tmin" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "tmin", period = "annual", trend = trn_tmin,
  title = "Overnight Minimum (tmin) — Annual Anomaly")
```

```{r plot-dtr, fig.cap="Diurnal temperature range (daytime maximum minus overnight minimum) annual mean for the Kootenay Lake Region. The downward trend indicates overnight lows are warming faster than daytime highs — the textbook day-night asymmetry shows up here.", echo = FALSE}
tmax_ts <- ts[ts$variable == "tmax" & ts$period == "annual", c("year", "value")]
tmin_ts <- ts[ts$variable == "tmin" & ts$period == "annual", c("year", "value")]
dtr <- merge(tmax_ts, tmin_ts, by = "year", suffixes = c("_max", "_min"))
dtr$dtr <- dtr$value_max - dtr$value_min

ggplot(dtr, aes(x = year, y = dtr)) +
  geom_line(color = "grey50") +
  geom_point(color = "grey30", size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "#b2182b", linewidth = 0.8) +
  labs(
    title = "Diurnal Temperature Range — Kootenay Lake Region",
    x = NULL,
    y = expression("Daytime maximum minus overnight minimum (" * degree * "C)")
  ) +
  theme_minimal(base_size = 12)
```

## Snowpack

Snowpack is the hinge of BC hydrology: winter precipitation falls as
snow, accumulates on the ground, and releases as meltwater across
spring and summer. That seasonal storage is the difference between a
late-summer creek that still flows and one that doesn't. It also
sets the timing of the spring freshet — the annual flood pulse that
shapes channel morphology, mobilizes spawning gravels, and refills
off-channel rearing habitat for resident salmonids. Cordillera-wide,
snowpack has been declining for decades [@mote_etal2018Dramaticdeclines;
@pederson_etal2011UnusualNature]. For four BC river basins (Fraser,
Peace, Columbia, Campbell), Najafi et al. [-@najafi_etal2017AttributionObserved]
attribute observed spring SWE decline to anthropogenic forcing —
including the Columbia basin, the parent system of the Kootenays.
For the neighbouring Fraser, Kang et al.
[-@kang_etal2016ImpactsRapidly] document a ~10-day advance of the
spring freshet over 1949-2006.

**For the Kootenay Lake Region the snowpack signal is sharper than
in northern BC reporting regions, in line with the Selkirks /
Purcells sitting at warmer latitudes than Peace-region snow zones.**
Annual snow water equivalent (SWE) is down **23%** (206 → 160 mm)
since the 1951–1980 reference. Annual snowfall and annual snowmelt
both fell about 15%, and the **snowmelt midpoint (DOY-50) shifted
12.6 days earlier** in the year. Annual peak SWE — the seasonal
maximum — dropped **92 mm** (-16%). And — distinct from the FWCP
Peace where total precipitation is roughly stable — **annual
precipitation in the Kootenay Lake Region has declined ~7%** since
the reference period, with a long-term Mann-Kendall p of 0.02.
Together this is a "warmer *and* drier" story rather than a
"warmer-only" one.

A few notes on how to read the snow numbers below.

ERA5-Land represents snow on a roughly 9 km grid — each grid cell is
a single number summarizing snow averaged over about 80 km² of mixed
terrain (saddles, slopes, valleys, forest, exposed alpine, all
combined). When we compare the model against a snow station inside
that cell, two distinct errors can stack: a **scale mismatch** (a
single point measurement isn't the same thing as an 80 km² average,
especially in mountain terrain where snow accumulation varies
sharply over short distances), and a **cell-mean bias** that the
model has at the Northern Hemisphere scale (ERA5-Land overestimates
mountain SWE by 150-200% even when compared against area-averaged
satellite estimates [@kouki_etal2023Evaluationsnow], traced to a
simplified snow layer in the underlying atmospheric model).

Our QA cross-check for this region used three usable BC ASWS
automated snow-pillow records inside the AOI (74 paired
station-years 1972-2025). Pooled correlation between ERA5-Land
peak SWE and station peak SWE is **r = 0.90** — meaningfully
better than the FWCP Peace's 0.51, meaning the model tracks
year-to-year variability well at these high-elevation Kootenay
sites. But the absolute bias is large and uniformly negative:
ERA5-Land is **40-63% too low** at Moyie Mountain (1835 m,
Purcells alpine) and Redfish Creek (2100 m, Selkirks alpine),
consistent with both stations sitting at high-snow microsites
within their 9 km cells. Per-site bias is approximately stable
over time at Moyie Mountain (regression p = 0.15) and marginal
at Redfish Creek (p = 0.07, slope of about -12 mm/yr widening),
so the trend-interpretability argument carries with one small
caveat: at the Redfish site specifically the model-station gap
appears to be widening slowly, which would weaken trend
defensibility there but not at the regional aggregate where many
sites contribute.

What does survive both kinds of error: **the gap between the model
and the stations is the same size in 2020 as it was in 1990.** The
bias is stable over time at every site (regression of
model-minus-station on year is non-significant, p > 0.2). That means
the *changes over time* shown below — peak snowpack dropping,
freshet shifting earlier — are real even though the absolute mm
numbers shouldn't be quoted as ground truth. For regional aggregates
(the numbers in this section are spatial means over hundreds of
cells), random point-vs-cell mismatches partially cancel, and stable
cell-mean biases preserve the trend signal.

The trend tests below use raw Mann-Kendall plus Theil-Sen — no
pre-whitening — which is the right call for our 76-year series with
strong climate trends per Yue and Wang
[-@yue_wang2002Applicabilityprewhitening] (pre-whitening
underestimates slope when a real trend exists).

### Seasonal snowpack curve

The four monthly-native snow variables — SWE, snowfall, snowmelt, and
snow cover — show *when* snow accumulates and melts. Aggregated to
the four standard meteorological seasons — winter (December–February),
spring (March–May), summer (June–August), and fall
(September–November) — the table below compares the recent decade
(2015-2025) against the pre-warming reference (1951-1980) directly.
The headline numbers above (summer SWE collapse, spring snowmelt rise)
are in the **summer** and **spring** rows.

```{r snow-seasonal-table, echo = FALSE}
snow_monthly <- c("swe", "snowfall", "snowmelt", "snow_cover")
season_order <- c("winter", "spring", "summer", "fall")

snow_seasonal <- cmp_pct[cmp_pct$variable %in% snow_monthly &
                          cmp_pct$period %in% season_order, ]
snow_seasonal$period <- factor(snow_seasonal$period, levels = season_order)
snow_seasonal$variable <- factor(snow_seasonal$variable, levels = snow_monthly,
                                  labels = c("SWE (mm)", "Snowfall (mm)",
                                             "Snowmelt (mm)", "Snow cover (%)"))
snow_seasonal <- snow_seasonal[order(snow_seasonal$variable, snow_seasonal$period), ]
snow_seasonal$mean_a <- round(snow_seasonal$mean_a, 1)
snow_seasonal$mean_b <- round(snow_seasonal$mean_b, 1)
snow_seasonal$difference <- round(snow_seasonal$difference, 1)
snow_seasonal <- snow_seasonal[, c("variable", "period", "mean_b",
                                    "mean_a", "difference")]
names(snow_seasonal) <- c("Variable", "Season",
                          "Pre-warming (1951–1980)",
                          "Recent (2015–2025)", "Δ %")

kableExtra::kable_styling(
  knitr::kable(snow_seasonal, label = NA,
    caption = "Seasonal snowpack: recent decade compared to pre-warming reference for the Kootenay Lake Region. Summer SWE collapse (-75%) and spring snowmelt rise (+37%) are the headline signals.",
    row.names = FALSE),
  bootstrap_options = c("striped", "hover", "condensed")
)
```

<br>

### Annual climate-departure signals

Four derived annual scalars capture the climate-departure signals the
literature treats as headline metrics for snow hydrology: peak
snowpack [@mote_etal2018Dramaticdeclines; @mote_etal2005DECLININGMOUNTAIN],
the date of melt midpoint
[@stewart_etal2005ChangesEarlier; @cayan_etal2001ChangesOnset],
freshet flashiness, and the
fraction of precipitation falling as snow [@knowles_etal2006TrendsSnowfall].
Two notes on our specific implementation. We use the actual annual
maximum of daily SWE rather than the April-1 SWE canon
[@pederson_etal2011UnusualNature] — equivalent in effect for BC
pixels (peak is at or near April 1) but date-insensitive. And our
freshet-flashiness metric (annual maximum of 7-day rolling daily
snowmelt) is upstream of the streamflow-based flashiness measures in
the literature [@stewart_etal2005ChangesEarlier;
@kang_etal2016ImpactsRapidly] — diagnostic of snowpack-side intensity
before routing through soil and channel storage.

```{r snow-swe-max, fig.cap="Annual peak snow water equivalent (swe_max) for the Kootenay Lake Region. ERA5-Land mm SWE (regional spatial mean)."}
trn_swe_max <- cd_trend(
  ano[ano$variable == "swe_max" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "swe_max", period = "annual",
                   trend = trn_swe_max,
                   title = "Annual peak SWE — Anomaly")
```

```{r snow-doy-50, fig.cap="Day of year when half the annual snowmelt has accumulated (snowmelt_doy_50). Earlier dates indicate an earlier freshet centroid."}
trn_doy <- cd_trend(
  ano[ano$variable == "snowmelt_doy_50" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "snowmelt_doy_50", period = "annual",
                   trend = trn_doy,
                   title = "Snowmelt 50% DOY — Anomaly")
```

```{r snow-rate-peak, fig.cap="Annual maximum of 7-day rolling daily snowmelt (snowmelt_rate_peak). Higher values indicate more concentrated freshet pulses."}
trn_rate <- cd_trend(
  ano[ano$variable == "snowmelt_rate_peak" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "snowmelt_rate_peak", period = "annual",
                   trend = trn_rate,
                   title = "Peak weekly melt rate — Anomaly")
```

```{r snow-fraction, fig.cap="Annual snowfall fraction (snowfall_fraction): the percent of annual precipitation that fell as snow. Anomaly is in percentage points."}
trn_frac <- cd_trend(
  ano[ano$variable == "snowfall_fraction" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "snowfall_fraction", period = "annual",
                   trend = trn_frac,
                   title = "Snowfall fraction — Anomaly")
```

### What this means for the Kootenay Lake Region

Three findings carry the snowpack story for the Kootenay Lake Region.

**Snow is leaving *and* falling less.** Annual snowfall is down 15%
(649 → 554 mm) — distinct from the FWCP Peace pattern just north,
where the snowpack decline was almost entirely about earlier melt
on roughly stable annual snowfall. The Kootenay Lake region is
warm enough at the relevant elevations that the snow-vs-rain
threshold is being crossed in the calendar — winter precipitation is
falling more often as rain instead of snow. This matches the
threshold finding from Knowles et al.
[-@knowles_etal2006TrendsSnowfall]: the strongest snowfall-fraction
declines in the western US occur where winter wet-day minimum
temperatures are warmer than -5 °C, which is the regime the
southern Kootenays sit in.

**The freshet is shifting into spring.** The day of year by which
half the year's snowmelt has happened (DOY-50) shifted **12.6 days
earlier** between the 1951-1980 reference and the recent decade —
in line with Stewart et al.'s [-@stewart_etal2005ChangesEarlier]
1-4 week earlier streamflow timing across western North America,
and with the ~10-day Fraser freshet advance documented by Kang et
al. [-@kang_etal2016ImpactsRapidly] for a basin whose southern
boundary is just north of the Kootenay Lake Region.

**Summer is becoming snow-free.** Summer SWE has collapsed by 73%
and summer snowmelt is down 61%. The high-elevation snowpack that
historically lingered into summer no longer does in the recent
decade. For aquatic ecosystems downstream, this is a loss of
late-season cold-water input to streams during the warmest, most
thermally stressful weeks of the year.

## Recent vs Pre-warming

The table below compares two windows directly — the recent decade
(2015–2025) against the pre-warming reference (1951–1980). Δ
absolute is the difference of the two means in the variable's native
units. Δ % is shown only for variables where it is meaningful
(precipitation, soil moisture, vapour pressure deficit, relative
humidity). The trend p column is the Mann-Kendall p-value of the
75-year (1951–present) trend on the same variable and period; it
tests for a steady year-on-year ramp, which is a related but distinct
question from "do these two windows differ".

The recent decade was 1.6 to 1.8 °C warmer than the pre-warming
reference for annual mean, daytime maximum, and overnight minimum,
with Mann-Kendall trend p-values below 0.001. Vapour pressure deficit
is up significantly. Annual precipitation was about 6 to 7 % *lower*
in the recent decade — and unlike the FWCP Peace just to the north,
the long-term Mann-Kendall trend test does confirm a steady
year-on-year decline (p ≈ 0.02). Soil moisture is roughly flat.
Relative humidity shows a small significant decline.

```{r compare-recipe, eval = FALSE}
cmp <- cd_compare(ts, window_a = 2015:2025, window_b = 1951:1980)
```

```{r compare-table, echo = FALSE}
trn_p <- trn[trn$trend_start == 1951, c("variable", "period", "mk_pvalue")]
names(trn_p)[3] <- "trend_p"
trn_p$trend_p <- round(trn_p$trend_p, 3)

no_pct_vars <- c(
  "tmean", "tmax", "tmin",
  # snow_cover and snowfall_fraction are already in % (pct_point_diff
  # anomaly type) — pct-of-baseline mixes units. snowmelt_doy_50 is a
  # day-of-year — pct-change of a date is meaningless.
  "snow_cover", "snowfall_fraction", "snowmelt_doy_50"
)
cmp_combined <- cmp |>
  dplyr::mutate(
    pct_change = ifelse(
      variable %in% no_pct_vars,
      NA_real_,
      round(100 * (mean_a - mean_b) / mean_b, 1)
    ),
    mean_a   = round(mean_a, 2),
    mean_b   = round(mean_b, 2),
    abs_diff = round(difference, 2)
  ) |>
  dplyr::select(variable, period, mean_a, mean_b, abs_diff, pct_change) |>
  merge(trn_p, by = c("variable", "period"), all.x = TRUE)

names(cmp_combined) <- c(
  "Variable", "Period",
  "Recent (2015–2025)", "Pre-warming (1951–1980)",
  "Δ absolute", "Δ %", "Trend p (75-yr)"
)

kableExtra::kable_styling(
  knitr::kable(cmp_combined, label = NA,
    caption = "Recent decade (2015-2025 mean) compared to pre-warming reference (1951-1980 mean) for the Kootenay Lake Region.",
    row.names = FALSE),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "360px")
```

<br>

## Spatial Pattern

The zonal mean reduces a region this size to a single number; the spatial
pattern carries the rest of the story.

Warming is not spatially uniform across the region. Total departures
range from about +1.2 °C at the lowest-warming pixels to +2.2 °C at
the highest, with a regional mean near +1.7 °C. Higher-elevation
pixels tend to show stronger warming — consistent with the
elevation-dependent warming signal documented at mid-latitude
mountain sites [@pepin_etal2015Elevationdependentwarming], though
not every mountain region shows the same pattern and the regional
evidence base remains heterogeneous [@rangwala_miller2012Climatechange].
The gradient here is mixed enough that no single axis (north-south
or east-west) carries the full pattern.

```{r spatial-tmean, fig.cap="Spatial pattern of annual mean temperature departure across the Kootenay Lake Region (2015-2025 mean minus 1951-1980 mean, degrees C).", fig.height=7, echo = FALSE}
# Pre-computed by data-raw/kootenay_lake_vignette_data.R. Live equivalent:
#   r_tmean <- cd_crop(catalog$href[catalog$variable == "tmean"
#                                   & catalog$period == "annual"], aoi)
#   years <- as.integer(names(r_tmean))
#   departure <- mean(r_tmean[[which(years >= 2015 & years <= 2025)]]) -
#                mean(r_tmean[[which(years >= 1951 & years <= 1980)]])
#   departure <- terra::mask(departure, aoi)
departure <- terra::rast(system.file(
  "vignette-data", "kootenay_lake_departure_tmean.tif", package = "cd"
))
names(departure) <- "Temperature departure"

ggplot() +
  geom_spatraster(data = departure) +
  geom_sf(data = ecoregions, fill = NA, color = "grey25",
          linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = aoi, fill = NA, color = "black", linewidth = 0.6) +
  geom_sf(data = lakes, fill = NA, color = "grey40", linewidth = 0.2) +
  # Highway: grey casing under warm yellow fill (gq reg_qgis_restoration)
  geom_sf(data = highways, color = "#9c9c9c", linewidth = 0.6) +
  geom_sf(data = highways, color = "#ffe585", linewidth = 0.4) +
  geom_sf(data = towns, color = "black", size = 2) +
  ggrepel::geom_label_repel(
    data = towns, aes(label = name, geometry = geom),
    stat = "sf_coordinates", size = 3, fill = "white", alpha = 0.8
  ) +
  scale_fill_distiller(
    palette = "RdBu", direction = -1,
    name = expression(Delta * degree * C)
  ) +
  coord_sf(
    xlim = c(aoi_bb["xmin"], aoi_bb["xmax"]),
    ylim = c(aoi_bb["ymin"], aoi_bb["ymax"])
  ) +
  labs(title = "Temperature Departure (2015-2025 vs 1951-1980)") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank())
```

## Per-Ecoregion Variation

The regional zonal mean averages over four ecoregions with different
elevations and exposures: Northern Columbia Mountains (NCM, the
dominant zone covering most of the AOI including the Selkirks and
Purcells), Selkirk-Bitterroot Foothills (SBF, the western lower-
elevation tier of LARL), Thompson-Okanagan Plateau (TOP, a small
sliver in LARL's far west), and Pacific and Cascade Ranges (PTR,
also a small sliver). To check whether the regional story holds
within each ecoregion — and where it does not — we run the same
pipeline on each ecoregion polygon individually.

All four ecoregions warmed at roughly the same rate. Precipitation
shows a region-wide decline; the magnitude is consistent across
ecoregions rather than concentrated in any one zone. Vapour
pressure deficit is up significantly across the region.

```{r per-ecoregion-recipe, eval = FALSE}
# For each ecoregion polygon, run the same cd pipeline as the regional one:
results <- lapply(seq_len(nrow(ecoregions)), function(i) {
  poly  <- ecoregions[i, ]
  ts_i  <- cd_extract(catalog, poly)
  bl_i  <- cd_baseline(ts_i, baseline_years = 1951:1980)
  ano_i <- cd_anomaly(ts_i, bl_i)
  trn_i <- cd_trend(ano_i, trend_start = c(1951, 1981))
  cmp_i <- cd_compare(ts_i, window_a = 2015:2025, window_b = 1951:1980)
  list(ano = ano_i, trn = trn_i, cmp = cmp_i)
})
```

```{r per-ecoregion-reshape, include = FALSE}
# Pre-computed equivalent loaded from the bundled rds; the reshape
# below builds long-form tables and trend-line segments for the
# per-ecoregion facet plots.
results <- lapply(vignette_data$ecoregion, function(x) {
  list(ano = x$ano, trn = x$trn, cmp = x$cmp_pct)
})
# Reorder to match ecoregions (already by area_km2 desc on disk)
results <- results[ecoregions$code]

# Long-form labels: "OMM: Omineca Mountains"
er_labels <- setNames(
  paste0(ecoregions$code, ": ",
         tools::toTitleCase(tolower(ecoregions$name))),
  ecoregions$code
)

ano_all <- do.call(rbind, lapply(seq_along(results), function(i) {
  o <- results[[i]]$ano
  o$ecoregion <- ecoregions$code[i]
  o
}))
ano_all$ecoregion <- factor(ano_all$ecoregion, levels = ecoregions$code)

# Theil-Sen trend segments for facet overlay — one row per
# (ecoregion x trend_start) combination so both 75-yr and 45-yr lines
# render in each facet.
build_segments <- function(var) {
  do.call(rbind, lapply(seq_along(results), function(i) {
    sub <- results[[i]]$trn
    sub <- sub[sub$variable == var & sub$period == "annual", ]
    if (nrow(sub) == 0) return(NULL)
    yrs <- ano_all$year[ano_all$variable == var & ano_all$ecoregion == ecoregions$code[i]]
    data.frame(
      ecoregion = factor(ecoregions$code[i], levels = ecoregions$code),
      window = paste0(sub$trend_start, "–", max(yrs)),
      x_start = sub$trend_start,
      x_end = max(yrs),
      y_start = sub$intercept + sub$slope * sub$trend_start,
      y_end = sub$intercept + sub$slope * max(yrs)
    )
  }))
}
trn_segs_tmean   <- build_segments("tmean")
trn_segs_prcp    <- build_segments("prcp")
trn_segs_swe_max <- build_segments("swe_max")
trn_segs_doy_50  <- build_segments("snowmelt_doy_50")
```

```{r facet-tmean, fig.cap="Annual mean temperature anomaly relative to the 1951-1980 baseline, by ecoregion. Dashed line is the 75-year Theil-Sen trend (1951-present); solid line is the 45-year trend (1981-present). A solid line steeper than the dashed line indicates accelerating warming.", fig.height=6, echo = FALSE}
ggplot(ano_all[ano_all$variable == "tmean" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_tmean,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = expression("Mean temperature anomaly (" * degree * C * ")")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r facet-prcp, fig.cap="Annual precipitation anomaly (% of the 1951-1980 baseline) by ecoregion. Dashed line is the 75-year trend (1951-present); solid line is the 45-year trend (1981-present). The two northernmost ecoregions (BMP, NRM) show statistically significant precipitation increases over the full record; the others do not.", fig.height=6, echo = FALSE}
ggplot(ano_all[ano_all$variable == "prcp" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_prcp,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#4575b4", `FALSE` = "#d73027")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "Precipitation anomaly (% of baseline)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

### Snow per ecoregion

Two snow metrics are worth viewing per-ecoregion: peak snow water
equivalent (annual snowpack magnitude) and DOY-50 (the day-of-year by
which half the year's snowmelt has already happened — earlier values
mean an earlier freshet). The two figures below show how each varies
ecoregion to ecoregion. Pair them with the Watershed Groups Across
Ecoregions section below to read each watershed group's dominant
ecoregion's signal.

```{r facet-swe-max, fig.cap="Annual peak snow water equivalent (SWE) anomaly by ecoregion, relative to the 1951-1980 baseline. Bars are mm of water-equivalent snowpack departure from baseline. Dashed line is the 75-year Theil-Sen trend (1951-present); solid line is the 45-year trend (1981-present).", fig.height=6, echo = FALSE}
ggplot(ano_all[ano_all$variable == "swe_max" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_swe_max,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#4575b4", `FALSE` = "#d73027")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "Peak SWE anomaly (mm)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r facet-doy-50, fig.cap="Snowmelt 50% day-of-year (DOY-50) anomaly by ecoregion, relative to the 1951-1980 baseline. Negative values (red) mean the freshet midpoint shifted earlier in the year. Dashed line is the 75-year Theil-Sen trend; solid line is the 45-year trend.", fig.height=6, echo = FALSE}
ggplot(ano_all[ano_all$variable == "snowmelt_doy_50" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_doy_50,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "DOY-50 anomaly (days; negative = earlier melt)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

### Snow per watershed group

The four watershed groups that make up the AOI map onto the FWCP
reporting unit directly. Per-WSG facet plots show the same
two-metric snow signal — peak SWE and DOY-50 — broken out by
watershed group rather than by ecoregion.

```{r per-wsg-recipe, eval = FALSE}
# Same cd pipeline applied per watershed group polygon:
wsg_results <- lapply(seq_len(nrow(wsgs)), function(i) {
  poly  <- wsgs[i, ]
  ts_i  <- cd_extract(catalog, poly)
  bl_i  <- cd_baseline(ts_i, baseline_years = 1951:1980)
  ano_i <- cd_anomaly(ts_i, bl_i)
  trn_i <- cd_trend(ano_i, trend_start = c(1951, 1981))
  list(ano = ano_i, trn = trn_i)
})
```

```{r per-wsg-reshape, include = FALSE}
wsg_results <- vignette_data$wsg
wsg_codes_v <- vignette_data$wsg_codes

ano_wsg <- do.call(rbind, lapply(seq_along(wsg_results), function(i) {
  o <- wsg_results[[i]]$ano
  o$wsg <- wsg_codes_v[i]
  o
}))
ano_wsg$wsg <- factor(ano_wsg$wsg, levels = wsg_codes_v)

build_segments_wsg <- function(var) {
  do.call(rbind, lapply(seq_along(wsg_results), function(i) {
    sub <- wsg_results[[i]]$trn
    sub <- sub[sub$variable == var & sub$period == "annual", ]
    if (nrow(sub) == 0) return(NULL)
    yrs <- ano_wsg$year[ano_wsg$variable == var & ano_wsg$wsg == wsg_codes_v[i]]
    data.frame(
      wsg = factor(wsg_codes_v[i], levels = wsg_codes_v),
      window = paste0(sub$trend_start, "–", max(yrs)),
      x_start = sub$trend_start,
      x_end = max(yrs),
      y_start = sub$intercept + sub$slope * sub$trend_start,
      y_end = sub$intercept + sub$slope * max(yrs)
    )
  }))
}
trn_wsg_swe_max <- build_segments_wsg("swe_max")
trn_wsg_doy_50  <- build_segments_wsg("snowmelt_doy_50")
```

```{r facet-wsg-swe-max, fig.cap="Annual peak SWE anomaly by watershed group. Bars are mm of water-equivalent snowpack departure from the 1951-1980 baseline. Dashed line is the 75-year Theil-Sen trend; solid line is the 45-year trend.", fig.height=5, echo = FALSE}
ggplot(ano_wsg[ano_wsg$variable == "swe_max" & ano_wsg$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_wsg_swe_max,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#4575b4", `FALSE` = "#d73027")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ wsg, ncol = 2) +
  labs(x = NULL, y = "Peak SWE anomaly (mm)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r facet-wsg-doy-50, fig.cap="Snowmelt DOY-50 anomaly by watershed group. Negative values (red) mean the freshet midpoint shifted earlier in the year.", fig.height=5, echo = FALSE}
ggplot(ano_wsg[ano_wsg$variable == "snowmelt_doy_50" & ano_wsg$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_wsg_doy_50,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ wsg, ncol = 2) +
  labs(x = NULL, y = "DOY-50 anomaly (days; negative = earlier melt)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r rollup, echo = FALSE}
get_75 <- function(trn, var) {
  trn[trn$variable == var & trn$period == "annual" & trn$trend_start == 1951, ]
}
rollup <- do.call(rbind, lapply(seq_along(results), function(i) {
  res <- results[[i]]
  cmp <- res$cmp
  tmean <- get_75(res$trn, "tmean")
  tmax  <- get_75(res$trn, "tmax")
  tmin  <- get_75(res$trn, "tmin")
  prcp  <- get_75(res$trn, "prcp")
  vpd   <- get_75(res$trn, "vpd")
  data.frame(
    Ecoregion = ecoregions$code[i],
    `tmean degC/dec` = round(10 * tmean$slope, 2),
    `tmax degC/dec`  = round(10 * tmax$slope, 2),
    `tmin degC/dec`  = round(10 * tmin$slope, 2),
    `prcp mm/yr` = round(prcp$slope, 3),
    `prcp p` = round(prcp$mk_pvalue, 3),
    `vpd hPa/dec` = round(10 * vpd$slope, 3),
    `vpd p` = round(vpd$mk_pvalue, 3),
    `prcp pct change` = round(
      cmp$difference[cmp$variable == "prcp" & cmp$period == "annual"], 1
    ),
    `soil moisture pct change` = round(
      cmp$difference[cmp$variable == "soil_moisture" & cmp$period == "annual"], 1
    ),
    check.names = FALSE
  )
}))
knitr::kable(rollup, row.names = FALSE,
  caption = "Per-ecoregion roll-up over the 75-year window (1951-present): annual mean, daytime maximum, and overnight minimum temperature trends (degrees C per decade); annual precipitation trend (mm per year) and Mann-Kendall p-value; annual vapour pressure deficit trend (hPa per decade) and p-value; and recent (2015-2025) vs pre-warming (1951-1980) percent change for precipitation and soil moisture. All temperature p-values are below 0.001. A tmin slope greater than the tmax slope indicates the textbook day-night asymmetry.")
```

Ecoregion codes used in the table above: `r paste0(ecoregions$code, " — ", ecoregions$name_tc, collapse = "; ")`.

<br>

## Watershed Groups Across Ecoregions

The Kootenay Lake Region is reported and managed at the watershed-group
scale — the British Columbia Freshwater Atlas hydrological reporting
unit, and the unit the Fish & Wildlife Compensation Program funds
work in. Four watershed groups make up the AOI for this
list. They span the four ecoregions in different ways. KOTL, DUNC,
and SLOC sit almost entirely within Northern Columbia Mountains
(NCM); LARL is the most-mixed group, splitting across NCM and the
Selkirk-Bitterroot Foothills (SBF). The map below shows the
watershed groups labelled with their codes, on top of ecoregion
fills. The table that follows gives the share of each watershed
group's area falling in each ecoregion.

```{r map-wsgs, fig.cap="The four watershed groups making up the AOI (KOTL, LARL, DUNC, SLOC) on top of ecoregion fills. Northern Columbia Mountains (NCM) covers most of the area; Selkirk-Bitterroot Foothills (SBF) covers the western portion of LARL; small slivers in the southwest belong to Thompson-Okanagan Plateau (TOP) and Pacific and Cascade Ranges (PTR).", fig.height=7, echo = FALSE}
wsgs <- st_read(ctx, layer = "wsgs", quiet = TRUE)

wsg_centroids <- suppressWarnings(st_centroid(wsgs))

ggplot() +
  geom_sf(data = ecoregions, aes(fill = name_tc), color = "grey45",
          linewidth = 0.3, alpha = 0.45) +
  geom_sf(data = aoi, fill = NA, color = "#2166ac", linewidth = 0.8) +
  geom_sf(data = wsgs, fill = NA, color = "grey25", linewidth = 0.4) +
  ggrepel::geom_label_repel(
    data = wsg_centroids,
    aes(label = code, geometry = geom),
    stat = "sf_coordinates",
    size = 2.6, fill = "white", alpha = 0.85,
    label.padding = unit(0.15, "lines"),
    min.segment.length = 0
  ) +
  scale_fill_brewer(palette = "Set3", name = "Ecoregion") +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_sf(
    xlim = c(aoi_bb["xmin"] - 0.4, aoi_bb["xmax"] + 0.4),
    ylim = c(aoi_bb["ymin"] - 0.5, aoi_bb["ymax"] + 0.3)
  ) +
  labs(title = "Watershed Groups on Ecoregion Backdrop") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.key.size = unit(0.4, "cm"))
```

```{r wsg-ecoregion-table, echo = FALSE}
# Pre-computed inline by data-raw/kootenay_lake_vignette_data.R
ovr <- vignette_data$wsg_eco_overlap
ovr_wide <- reshape(
  ovr[, c("wsg_code", "ecoregion_code", "pct_of_wsg")],
  idvar = "wsg_code", timevar = "ecoregion_code", direction = "wide"
)
names(ovr_wide) <- gsub("^pct_of_wsg\\.", "", names(ovr_wide))
# Order WSG rows alphabetically; replace NAs with 0
ovr_wide <- ovr_wide[order(ovr_wide$wsg_code), ]
for (col in setdiff(names(ovr_wide), "wsg_code")) {
  ovr_wide[[col]][is.na(ovr_wide[[col]])] <- 0
}
# Add full WSG name
wsg_names <- setNames(ovr$wsg_name, ovr$wsg_code)[!duplicated(ovr$wsg_code)]
ovr_wide$Name <- wsg_names[ovr_wide$wsg_code]
ovr_wide <- ovr_wide[, c("wsg_code", "Name",
                          intersect(c("NCM","SBF","TOP","PTR"), names(ovr_wide)))]
names(ovr_wide)[1] <- "Code"
for (col in setdiff(names(ovr_wide), c("Code", "Name"))) {
  names(ovr_wide)[names(ovr_wide) == col] <- paste0(col, " %")
}

kableExtra::kable_styling(
  knitr::kable(ovr_wide, label = NA,
    caption = "Watershed groups in the Kootenay Lake Region with the percent of each group's area falling in each ecoregion. KOTL, DUNC, and SLOC are essentially within Northern Columbia Mountains; LARL is the only WSG that meaningfully splits across two ecoregions (NCM + SBF), making it the place where ecoregion-level findings can pull in two directions.",
    row.names = FALSE),
  bootstrap_options = c("striped", "hover", "condensed")
)
```

<br>

## Interpretation

Three findings emerge from the climate departures shown above.

**Warming is broad and fast.** All four ecoregions warmed between
+1.6 °C and +1.9 °C cumulative since 1951, at a rate near
+0.25 °C per decade, with Mann-Kendall p-values below 0.001 in every
ecoregion. The trend is statistically significant beyond reasonable
doubt of being random noise. Daily maximum, daily minimum, and daily
mean temperatures all tell the same story. The whole region moved
together — there is no warming hot-spot at the ecoregion scale.

**Precipitation is declining and the decline is regionally
consistent.** This is the place where the Kootenay Lake Region
diverges sharpest from the FWCP Peace just north. The Peace
showed mostly-flat annual precipitation with significant
*increases* in two of its five ecoregions; the Kootenays show a
**~7% decline** in annual precipitation from the 1951-1980
reference to the 2015-2025 recent decade, with a long-term
Mann-Kendall p of 0.02. Spatially the decline is consistent across
all four ecoregions of the AOI, so the pattern at the regional
average mirrors what each watershed group sees individually. All
four watershed groups in the AOI are sitting in the same drying
signal.

**The atmosphere is drying.** Vapour pressure deficit — the gap
between how much water the air could hold and how much it actually
does — is up significantly across the region, mirroring the
continental-scale drying that @ficklin_novick2017Historicprojected
documented for the United States as a whole, with the strongest
historical VPD increases concentrated in the western and southern
portions, driven by combined air-temperature increases and
relative-humidity declines. Combined
with declining precipitation, this is a "double-dipping" signal:
less water arriving as precipitation, *and* warmer air pulling more
water out of soil and vegetation through evapotranspiration. Soil
moisture is essentially flat despite the precipitation decline —
the warmer atmosphere is drinking surface moisture at a rate that
keeps the topsoil layer roughly even, but the resulting
late-summer water deficit shows up in stream baseflow rather than
in soil-moisture statistics.

**Snow is leaving *and* falling less.** This is the second place
the Kootenays diverge from the FWCP Peace pattern. The Peace's
snowpack decline was almost entirely about earlier melt on
roughly stable annual snowfall. The Kootenays show **annual
snowfall down 15%** alongside the freshet shift — winter
precipitation is falling more often as rain instead of snow,
matching the threshold finding from Knowles et al.
[-@knowles_etal2006TrendsSnowfall] for sites with winter wet-day
minimum temperatures warmer than -5 °C. Annual snow water
equivalent (SWE) is down 23% (206 → 160 mm); summer SWE has
collapsed by 73%; the snowmelt 50% day-of-year shifted **12.6
days earlier** between the reference and recent windows. The
per-WSG facet plot above shows the freshet-timing shift clearly
in all four watershed groups; the per-ecoregion breakdown shows
the signal is not concentrated in any one ecoregion.

For the cold-water resident salmonids the Kootenay Lake region
supports — bull trout, Gerrard rainbow trout, mountain whitefish,
kokanee — these signals compound. Stream temperatures are likely
rising in step with warmer ambient air temperatures.
@mantua_etal2010Climatechange modelled this for Washington State
watersheds and found that warming summer stream temperatures
combined with altered low flows would reduce reproductive success
for many salmon populations, and @eaton_scheller1996Effectsclimate
showed across 57 US fish species that cold- and cool-water species
lose substantially more thermal habitat than warm-water species
under the same forcing. Both findings carry directly into the BC
interior cold-water cohort the Kootenay Lake Region supports. The evapotranspiration
imbalance means low-flow conditions in late summer are not being
relieved (precipitation is falling, not rising as in the Peace);
the cold-water input that high-elevation snowpack provides to
streams during the warmest, most thermally stressful weeks of
summer is dropping in parallel with summer SWE; and the spring
freshet — the dominant high-flow event that shapes channel
morphology, mobilizes spawning gravels, and refills off-channel
rearing habitat — is shifting weeks earlier. The neighbouring
Fraser Basin shows the same shift —
@kang_etal2016ImpactsRapidly documented a ~10-day advance in the
spring freshet at Hope between 1949 and 2006, with persistent
declines through autumn recession just when salmon are migrating
up the Fraser. Lower
Columbia River reaches below Hugh Keenleyside Dam are
dam-fragmented and not anadromous, so the ecological framing is
about resident salmonids and their habitat rather than salmon
migrations.

## References
