---
title: "Climate Departure for the Kootenay Lake Region"
output:
  bookdown::html_document2:
    number_sections: false
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Climate Departure for the Kootenay Lake Region}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
After adding or removing pandoc citation markers in this vignette
(or in peace-fwcp.Rmd), regenerate vignettes/references.bib by
running:

  Rscript data-raw/regenerate_bib.R

Pulls source records from Zotero via Better BibTeX (rbbt -> BBT).
Requires Zotero running with the BBT plugin enabled. CI does not
run this script — the bib is committed and pkgdown reads it.
-->


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r libraries, include = FALSE}
library(terra)
library(ggplot2)
library(tidyterra)
```

The `cd` package builds on ERA5-Land hourly reanalysis (1950–present,
~9 km native grid) fetched from the DestinE Earth Data Hub, aggregated
to monthly, seasonal, and annual periods over British Columbia, and
published as Cloud-Optimized GeoTIFFs alongside a static STAC catalog
in a public S3 bucket. R consumer functions read those GeoTIFFs
directly via GDAL, crop to any area of interest, and compute baselines,
anomalies, and Mann-Kendall / Theil-Sen trend statistics.

This vignette runs the consumer pipeline on a four-watershed-group
area in the southern interior of British Columbia: Kootenay Lake
(`KOTL`), Lower Arrow Lake (`LARL`, which covers Trail / Rossland /
Red Mountain at its south end), Duncan Lake (`DUNC`, draining the
north end of Kootenay Lake including the Lardeau valley), and Slocan
River (`SLOC`, between the Selkirks and Monashees). The four
watershed groups together total ~24,000 km² and span the east-west
precipitation gradient that defines the snowpack story for the
region — Selkirk Pacific spillover west of Kootenay Lake, Purcell
rain shadow east.

## Area of Interest

The bundled area of interest is the union of the four watershed
groups (single multi-polygon, EPSG:4326). Any `sf` polygon works. The
watersheds in this region all drain to the Columbia River — Kootenay
Lake feeds the Kootenay River which joins the Columbia at Castlegar;
the Slocan and Lower Arrow drain directly into the Columbia.

We also carry British Columbia ecoregions through the analysis.
Ecoregions partition the province into broad climate-physiography
zones based on latitude, elevation, and dominant vegetation. Climate
departure can vary across them in ways the regional average hides,
and we use them later as the sub-region for the per-ecoregion
breakdown.

```{r load-aoi, echo = FALSE}
library(cd)
library(sf)

aoi <- st_read(
  system.file("extdata", "example_aoi_kootenay_lake.gpkg", package = "cd"),
  quiet = TRUE
)

area_km2 <- as.numeric(sum(st_area(st_transform(aoi, 3005)))) / 1e6
```

```{r load-context, include = FALSE}
ctx <- system.file("extdata", "context_kootenay_lake.gpkg", package = "cd")
towns       <- st_read(ctx, layer = "towns",      quiet = TRUE)
lakes       <- st_read(ctx, layer = "lakes",      quiet = TRUE)
rivers      <- st_read(ctx, layer = "rivers",     quiet = TRUE)
streams     <- st_read(ctx, layer = "streams",    quiet = TRUE)
highways    <- st_read(ctx, layer = "highways",   quiet = TRUE)
ecoregions  <- st_read(ctx, layer = "ecoregions", quiet = TRUE)
ecoregions  <- ecoregions[ecoregions$area_km2 > 50, ]
ecoregions$name_tc <- tools::toTitleCase(tolower(ecoregions$name))

wsgs <- st_read(ctx, layer = "wsgs", quiet = TRUE)
```

```{r map-location, fig.cap="Kootenay Lake Region area of interest (~24,000 km²) in southern interior British Columbia, coloured by ecoregion. Kootenay Lake dominates the basin; Lower Arrow Lake and the Slocan are visible to the west.", fig.height=7, echo = FALSE}
aoi_bb <- st_bbox(aoi)

ggplot() +
  geom_sf(data = ecoregions, aes(fill = name_tc), color = "grey45",
          linewidth = 0.3, alpha = 0.45) +
  geom_sf(data = aoi, fill = NA, color = "#2166ac", linewidth = 0.8) +
  geom_sf(data = lakes, fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = rivers, fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = streams, color = "#74a9cf", linewidth = 0.3, alpha = 0.6) +
  # Highway: grey casing under warm yellow fill (gq reg_qgis_restoration)
  geom_sf(data = highways, color = "#9c9c9c", linewidth = 0.75) +
  geom_sf(data = highways, color = "#ffe585", linewidth = 0.5) +
  geom_sf(data = towns, color = "black", size = 2.5) +
  ggrepel::geom_label_repel(
    data = towns,
    aes(label = name, geometry = geom),
    stat = "sf_coordinates",
    size = 3.2, fill = "white", alpha = 0.85,
    label.padding = unit(0.2, "lines"),
    min.segment.length = 0
  ) +
  scale_fill_brewer(palette = "Set3", name = "Ecoregion") +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_sf(
    xlim = c(aoi_bb["xmin"] - 0.4, aoi_bb["xmax"] + 0.4),
    ylim = c(aoi_bb["ymin"] - 0.5, aoi_bb["ymax"] + 0.3)
  ) +
  labs(title = "Kootenay Lake Region",
       subtitle = paste0(round(area_km2), " km² — ", nrow(ecoregions), " ecoregions")) +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.key.size = unit(0.4, "cm"))
```

## Connect to the Data Catalog

The producer pipeline in this repository fetches ERA5-Land hourly
reanalysis data, derives variables (vapour pressure deficit, relative
humidity, soil moisture), aggregates to monthly / seasonal / annual,
and writes Cloud-Optimized GeoTIFFs to S3. Alongside the GeoTIFFs we
publish a static SpatioTemporal Asset Catalog (STAC) — a small set of
JSON files that index the assets. Both the GeoTIFFs and the catalog
JSON live at
<https://stac-era5-land.s3.us-west-2.amazonaws.com/catalog.json>.

`cd::cd_catalog()` reads that URL by default. The Cloud-Optimized GeoTIFFs
are also usable directly outside R — for example, in QGIS via the STAC
plugin, in `gdalcubes`, or with any STAC-aware client.

```{r catalog}
catalog <- cd::cd_catalog()
kableExtra::kable_styling(
  knitr::kable(catalog, label = NA,
    caption = "Available climate variables and periods in the STAC catalog."),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "320px")
```

<br>

## Extract Climate Time Series

`cd::cd_extract()` crops each cloud-hosted GeoTIFF to the area of interest
and computes the spatial mean per year. For an area of interest of
this size a live extraction takes a few seconds per variable. To keep
the vignette fast and reproducible we load a pre-computed result of
exactly that call.

```{r extract-recipe, eval = FALSE}
ts <- cd::cd_extract(catalog, aoi)
head(ts, 10)
```

```{r extract-load, echo = FALSE}
# Load the pre-computed equivalent so the vignette renders without
# hitting the network. ts is the time series tibble; bl_early, ano,
# trn, cmp, and cmp_pct are the downstream products of the cd
# pipeline (computed once in data-raw/kootenay_lake_vignette_data.R).
vignette_data <- readRDS(system.file(
  "vignette-data", "kootenay_lake.rds", package = "cd"
))
ts       <- vignette_data$regional$ts
bl_early <- vignette_data$regional$bl
ano      <- vignette_data$regional$ano
trn      <- vignette_data$regional$trn
cmp      <- vignette_data$regional$cmp
cmp_pct  <- vignette_data$regional$cmp_pct

knitr::kable(head(ts, 10),
  caption = "First 10 rows of the extracted climate time series.")
```

## Trends

An anomaly is the difference between a year's value and the average value
over a reference period. Here the reference is **1951–1980** — the three
decades before climate change accelerated, the same baseline
@hansen_etal2012Perceptionclimate used to track global temperature
outliers. Saying a year is "+1.5 °C" means it was 1.5 °C warmer than the
average year in 1951–1980.

The trend table shows two rows per variable, fit from two different start
years:

- **1951–present (75 years)** — the long view. How much has the variable
  changed since the pre-warming reference?
- **1981–present (45 years)** — the recent view. Starts at the beginning
  of the most recent WMO "climate normal" (1981–2010), the period used in
  IPCC reports, so the numbers are easy to compare against published
  summaries.

If the 45-year slope is **steeper** than the 75-year slope, change has
accelerated. **Shallower** means it has slowed. **Similar** means a
roughly steady rate across the full record. "Total Change" is the slope
times the number of years in the window — the cumulative shift.

```{r trend-recipe, eval = FALSE}
bl_early <- cd::cd_baseline(ts, baseline_years = 1951:1980)
ano      <- cd::cd_anomaly(ts, bl_early)
trn      <- cd::cd_trend(ano, trend_start = c(1951, 1981))
cd::cd_summary(trn)
```

```{r trend-table, echo = FALSE}
kableExtra::kable_styling(
  knitr::kable(cd::cd_summary(trn), label = NA,
    caption = "Trend statistics for all variables and periods, Kootenay Lake Region."),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "320px")
```

<br>

```{r plot-prcp, fig.cap="Annual precipitation anomaly (% of 1951-1980 baseline) for the Kootenay Lake Region. The per-ecoregion mean temperature panel is shown later in the Per-Ecoregion section."}
cd::cd_plot_timeseries(
  ano, variable = "prcp", period = "annual", trend = trn,
  title = "Annual Precipitation Anomaly — Kootenay Lake Region"
)
```

## Recent Decade vs Pre-Warming Reference

The table below compares the recent decade (2015–2025) directly
against the pre-warming reference (1951–1980). **Δ absolute** is the
difference of the two means in the variable's native units.
**Δ %** is shown only where percent change makes physical sense
(precipitation, soil moisture, VPD, relative humidity).

Two p-values appear side-by-side because they answer different
questions. **Δ p (windows)** asks whether the recent decade differs
from the pre-warming reference. **Trend p (75-yr)** asks whether the
full series ramps year-on-year. A sudden step change shows up as
significant Δ p but not trend p; a gradual ramp shows up as both.

The recent decade was 1.6 to 1.8 °C warmer than the pre-warming
reference for annual mean, daytime maximum, and overnight minimum,
with both window and trend p-values below 0.001. Vapour pressure
deficit is up significantly on both tests. Annual precipitation
was about 6 to 7 % *lower* in the recent decade, and the long-term
Mann-Kendall trend test confirms a steady year-on-year decline
(trend p ≈ 0.02). Soil moisture is roughly flat. Relative humidity
shows a small significant decline.

```{r compare-recipe, eval = FALSE}
cmp <- cd::cd_compare(ts)   # defaults: 2015–2025 vs 1951–1980, Welch t-test
```

```{r compare-table, echo = FALSE}
# Recompute cmp from the cached ts so the p_value column reflects the
# current cd_compare() defaults — the precomputed rds was generated
# before #43 added the test argument.
cmp <- cd::cd_compare(ts)
trn_p <- trn[trn$trend_start == 1951, c("variable", "period", "mk_pvalue")]
names(trn_p)[3] <- "trend_p"
trn_p$trend_p <- round(trn_p$trend_p, 3)

no_pct_vars <- c(
  "tmean", "tmax", "tmin",
  # snow_cover and snowfall_fraction are already in % (pct_point_diff
  # anomaly type) — pct-of-baseline mixes units. snowmelt_doy_50 is a
  # day-of-year — pct-change of a date is meaningless.
  "snow_cover", "snowfall_fraction", "snowmelt_doy_50"
)
cmp_combined <- cmp |>
  dplyr::mutate(
    pct_change = ifelse(
      variable %in% no_pct_vars,
      NA_real_,
      round(100 * (mean_a - mean_b) / mean_b, 1)
    ),
    mean_a    = round(mean_a, 2),
    mean_b    = round(mean_b, 2),
    abs_diff  = round(difference, 2),
    window_p  = round(p_value, 3)
  ) |>
  dplyr::select(variable, period, mean_a, mean_b, abs_diff, pct_change, window_p) |>
  merge(trn_p, by = c("variable", "period"), all.x = TRUE)

names(cmp_combined) <- c(
  "Variable", "Period",
  "Recent (2015–2025)", "Pre-warming (1951–1980)",
  "Δ absolute", "Δ %", "Δ p (windows)", "Trend p (75-yr)"
)

kableExtra::kable_styling(
  knitr::kable(cmp_combined, label = NA,
    caption = "Recent decade (2015-2025 mean) compared to pre-warming reference (1951-1980 mean) for the Kootenay Lake Region.",
    row.names = FALSE),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "360px")
```

<br>

## Daytime Highs and Overnight Lows

Alongside the daily mean, the cd package ships **tmax** (daytime
maximum) and **tmin** (overnight minimum). When overnight minimums
warm faster than daytime maximums, that's the "day-night
asymmetry" — a textbook fingerprint of greenhouse warming, first
documented globally by @karl_etal1993NewPerspective (overnight
minimums rose roughly three times faster than daytime maximums
between 1951 and 1990). Whether a region shows the signal depends
on local geography (valley inversions, snow cover, slope-aspect mix).

**For the Kootenay Lake Region, overnight minimums are warming
faster than daytime maximums.** Daytime maximums warmed about
+0.024 °C/year since 1951 (+1.8 °C cumulative); overnight minimums
warmed about +0.027 °C/year (+2.0 °C cumulative). The overnight
side has gained roughly 0.2 °C more than the daytime side over the
full record.

```{r plot-tmaxmin, fig.cap="Annual tmax (top) and tmin (bottom) anomalies for the Kootenay Lake Region, relative to the 1951-1980 baseline. Dashed line is the 75-year Theil-Sen trend (1951-present); solid line is the 45-year trend (1981-present).", echo = FALSE, fig.height=6}
ano_dn <- ano[ano$variable %in% c("tmax", "tmin") & ano$period == "annual", ]
ano_dn$variable <- factor(ano_dn$variable, levels = c("tmax", "tmin"),
                          labels = c("Daytime maximum (tmax)",
                                     "Overnight minimum (tmin)"))
trn_dn <- cd::cd_trend(
  ano[ano$variable %in% c("tmax", "tmin") & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
trn_dn$variable <- factor(trn_dn$variable, levels = c("tmax", "tmin"),
                          labels = c("Daytime maximum (tmax)",
                                     "Overnight minimum (tmin)"))
trn_dn_lines <- do.call(rbind, lapply(seq_len(nrow(trn_dn)), function(i) {
  yrs <- ano_dn$year[ano_dn$variable == trn_dn$variable[i]]
  data.frame(
    variable = trn_dn$variable[i],
    window = paste0(trn_dn$trend_start[i], "–", max(yrs)),
    x_start = trn_dn$trend_start[i],
    x_end = max(yrs),
    y_start = trn_dn$intercept[i] + trn_dn$slope[i] * trn_dn$trend_start[i],
    y_end = trn_dn$intercept[i] + trn_dn$slope[i] * max(yrs)
  )
}))

ggplot(ano_dn, aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_dn_lines,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ variable, ncol = 1) +
  labs(x = NULL, y = expression("Anomaly (" * degree * C * ")")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Snowpack

In BC, most of the year's runoff starts as winter snow. The snowpack
stores it, then releases it as meltwater across spring and summer.
That delayed release is what keeps creeks flowing through late
summer and what drives the spring freshet — the annual flood pulse
that shapes channels, mobilizes spawning gravels, and refills
off-channel rearing habitat for resident salmonids.

Cordillera-wide, snowpack has been declining for decades
[@mote_etal2018Dramaticdeclines; @pederson_etal2011UnusualNature].
Najafi et al. [-@najafi_etal2017AttributionObserved] attribute the
observed spring SWE decline in the Columbia basin — the parent
system of the Kootenays — to human-caused warming.

**The Kootenay Lake Region snowpack signal is sharp.**
Annual snow water equivalent (SWE) is down **23%** (206 → 160 mm)
since the 1951–1980 reference. Annual snowfall and annual snowmelt
both fell about 15%, and the **snowmelt midpoint (DOY-50) shifted
12.6 days earlier** in the year. Annual peak SWE — the seasonal
maximum — dropped **92 mm** (-16%). **Annual precipitation in the
Kootenay Lake Region has declined ~7%** since the reference period,
with a long-term Mann-Kendall p of 0.02. Together this is a
"warmer *and* drier" story rather than a "warmer-only" one.

A note on how to read the snow numbers below. ERA5-Land works on a
roughly 9 km grid: each cell is a single number averaging snow over
about 80 km² of mixed terrain. Cross-checking against three BC
automated snow stations inside the AOI (74 paired station-years,
1972–2025), the model is **40–63% too low on absolute peak SWE**,
but the **bias is stable over time** — the gap between model and
stations in 2020 is the same size as in 1990 — and **year-to-year
correlation is strong** (r = 0.90). So the *trends and changes*
below are credible. The absolute mm values are not ground truth.

### Seasonal snowpack curve

The four monthly-native snow variables — SWE, snowfall, snowmelt, and
snow cover — show *when* snow accumulates and melts. Aggregated to
the four standard meteorological seasons — winter (December–February),
spring (March–May), summer (June–August), and fall
(September–November) — the table below compares the recent decade
(2015-2025) against the pre-warming reference (1951-1980) directly.
The headline numbers above (summer SWE collapse, spring snowmelt rise)
are in the **summer** and **spring** rows.

```{r snow-seasonal-table, echo = FALSE}
snow_monthly <- c("swe", "snowfall", "snowmelt", "snow_cover")
season_order <- c("winter", "spring", "summer", "fall")

snow_seasonal <- cmp_pct[cmp_pct$variable %in% snow_monthly &
                          cmp_pct$period %in% season_order, ]
snow_seasonal$period <- factor(snow_seasonal$period, levels = season_order)
snow_seasonal$variable <- factor(snow_seasonal$variable, levels = snow_monthly,
                                  labels = c("SWE (mm)", "Snowfall (mm)",
                                             "Snowmelt (mm)", "Snow cover (%)"))
snow_seasonal <- snow_seasonal[order(snow_seasonal$variable, snow_seasonal$period), ]
snow_seasonal$mean_a <- round(snow_seasonal$mean_a, 1)
snow_seasonal$mean_b <- round(snow_seasonal$mean_b, 1)
snow_seasonal$difference <- round(snow_seasonal$difference, 1)
snow_seasonal <- snow_seasonal[, c("variable", "period", "mean_b",
                                    "mean_a", "difference")]
names(snow_seasonal) <- c("Variable", "Season",
                          "Pre-warming (1951–1980)",
                          "Recent (2015–2025)", "Δ %")

kableExtra::kable_styling(
  knitr::kable(snow_seasonal, label = NA,
    caption = "Seasonal snowpack: recent decade compared to pre-warming reference for the Kootenay Lake Region. Summer SWE collapse (-75%) and spring snowmelt rise (+37%) are the headline signals.",
    row.names = FALSE),
  bootstrap_options = c("striped", "hover", "condensed")
)
```

<br>

### Annual snowpack signals

Three annual numbers capture how the snowpack is changing:

- **Peak SWE** — the year's maximum snow water equivalent. How much
  snow is in the bank at the height of winter?
  [@mote_etal2018Dramaticdeclines; @mote_etal2005DECLININGMOUNTAIN]
- **Snowmelt 50% day-of-year (DOY-50)** — the day by which half the
  year's melt has run off. Earlier = freshet shifting into spring.
  [@stewart_etal2005ChangesEarlier; @cayan_etal2001ChangesOnset]
- **Snowfall fraction** — the share of annual precipitation that
  falls as snow rather than rain.
  [@knowles_etal2006TrendsSnowfall]

```{r snow-annual, fig.cap="Annual snowpack signals for the Kootenay Lake Region, relative to the 1951-1980 baseline. Each panel uses its own y-axis. Dashed line is the 75-year Theil-Sen trend; solid line is the 45-year trend.", echo = FALSE, fig.height=7}
snow_vars <- c("swe_max", "snowmelt_doy_50", "snowfall_fraction")
snow_labels <- c("Peak SWE (mm)",
                 "Snowmelt 50% DOY (days)",
                 "Snowfall fraction (% points)")
ano_snow <- ano[ano$variable %in% snow_vars & ano$period == "annual", ]
ano_snow$variable <- factor(ano_snow$variable, levels = snow_vars,
                            labels = snow_labels)
trn_snow <- cd::cd_trend(
  ano[ano$variable %in% snow_vars & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
trn_snow$variable <- factor(trn_snow$variable, levels = snow_vars,
                            labels = snow_labels)
trn_snow_lines <- do.call(rbind, lapply(seq_len(nrow(trn_snow)), function(i) {
  yrs <- ano_snow$year[ano_snow$variable == trn_snow$variable[i]]
  data.frame(
    variable = trn_snow$variable[i],
    window = paste0(trn_snow$trend_start[i], "–", max(yrs)),
    x_start = trn_snow$trend_start[i],
    x_end = max(yrs),
    y_start = trn_snow$intercept[i] + trn_snow$slope[i] * trn_snow$trend_start[i],
    y_end = trn_snow$intercept[i] + trn_snow$slope[i] * max(yrs)
  )
}))

ggplot(ano_snow, aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_snow_lines,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ variable, ncol = 1, scales = "free_y") +
  labs(x = NULL, y = "Annual anomaly") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### What this means for the Kootenay Lake Region

Three findings carry the snowpack story for the Kootenay Lake Region.

**Snow is leaving *and* falling less.** Annual snowfall is down 15%
(649 → 554 mm). The Kootenay Lake region is warm enough at the
relevant elevations that the snow-vs-rain threshold is being crossed
in the calendar — winter precipitation is falling more often as rain
instead of snow. This matches the
threshold finding from Knowles et al.
[-@knowles_etal2006TrendsSnowfall]: the strongest snowfall-fraction
declines in the western US occur where winter wet-day minimum
temperatures are warmer than -5 °C, which is the regime the
southern Kootenays sit in.

**The freshet is shifting into spring.** The day of year by which
half the year's snowmelt has happened (DOY-50) shifted **12.6 days
earlier** between the 1951-1980 reference and the recent decade —
in line with Stewart et al.'s [-@stewart_etal2005ChangesEarlier]
1-4 week earlier streamflow timing across western North America,
and with the ~10-day Fraser freshet advance documented by Kang et
al. [-@kang_etal2016ImpactsRapidly] for a basin whose southern
boundary is just north of the Kootenay Lake Region.

**Summer is becoming snow-free.** Summer SWE has collapsed by 73%
and summer snowmelt is down 61%. The high-elevation snowpack that
historically lingered into summer no longer does in the recent
decade. For aquatic ecosystems downstream, this is a loss of
late-season cold-water input to streams during the warmest, most
thermally stressful weeks of the year.

## Spatial Pattern

The zonal mean reduces a region this size to a single number; the spatial
pattern carries the rest of the story.

Warming is not spatially uniform across the region. Total departures
range from about +1.2 °C at the lowest-warming pixels to +2.2 °C at
the highest, with a regional mean near +1.7 °C. Higher-elevation
pixels tend to show stronger warming — consistent with the
elevation-dependent warming signal documented at mid-latitude
mountain sites [@pepin_etal2015Elevationdependentwarming], though
not every mountain region shows the same pattern and the regional
evidence base remains heterogeneous [@rangwala_miller2012Climatechange].
The gradient here is mixed enough that no single axis (north-south
or east-west) carries the full pattern.

```{r spatial-tmean, fig.cap="Spatial pattern of annual mean temperature departure across the Kootenay Lake Region (2015-2025 mean minus 1951-1980 mean, degrees C).", fig.height=7, echo = FALSE}
# Pre-computed by data-raw/kootenay_lake_vignette_data.R. Live equivalent:
#   r_tmean <- cd::cd_crop(catalog$href[catalog$variable == "tmean"
#                                   & catalog$period == "annual"], aoi)
#   years <- as.integer(names(r_tmean))
#   departure <- mean(r_tmean[[which(years >= 2015 & years <= 2025)]]) -
#                mean(r_tmean[[which(years >= 1951 & years <= 1980)]])
#   departure <- terra::mask(departure, aoi)
departure <- terra::rast(system.file(
  "vignette-data", "kootenay_lake_departure_tmean.tif", package = "cd"
))
names(departure) <- "Temperature departure"

ggplot() +
  geom_spatraster(data = departure) +
  geom_sf(data = ecoregions, fill = NA, color = "grey25",
          linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = aoi, fill = NA, color = "black", linewidth = 0.6) +
  geom_sf(data = lakes, fill = NA, color = "grey40", linewidth = 0.2) +
  # Highway: grey casing under warm yellow fill (gq reg_qgis_restoration)
  geom_sf(data = highways, color = "#9c9c9c", linewidth = 0.6) +
  geom_sf(data = highways, color = "#ffe585", linewidth = 0.4) +
  geom_sf(data = towns, color = "black", size = 2) +
  ggrepel::geom_label_repel(
    data = towns, aes(label = name, geometry = geom),
    stat = "sf_coordinates", size = 3, fill = "white", alpha = 0.8
  ) +
  scale_fill_distiller(
    palette = "RdBu", direction = -1,
    name = expression(Delta * degree * C)
  ) +
  coord_sf(
    xlim = c(aoi_bb["xmin"], aoi_bb["xmax"]),
    ylim = c(aoi_bb["ymin"], aoi_bb["ymax"])
  ) +
  labs(title = "Temperature Departure (2015-2025 vs 1951-1980)") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank())
```

## Per-Ecoregion Variation

The regional zonal mean averages over four ecoregions with different
elevations and exposures: Northern Columbia Mountains (NCM, the
dominant zone covering most of the AOI including the Selkirks and
Purcells), Selkirk-Bitterroot Foothills (SBF, the western lower-
elevation tier of LARL), Thompson-Okanagan Plateau (TOP, a small
sliver in LARL's far west), and Pacific and Cascade Ranges (PTR,
also a small sliver). To check whether the regional story holds
within each ecoregion — and where it does not — we run the same
pipeline on each ecoregion polygon individually.

All four ecoregions warmed at roughly the same rate. Precipitation
shows a region-wide decline; the magnitude is consistent across
ecoregions rather than concentrated in any one zone. Vapour
pressure deficit is up significantly across the region.

```{r per-ecoregion-recipe, eval = FALSE}
# For each ecoregion polygon, run the same cd pipeline as the regional one:
results <- lapply(seq_len(nrow(ecoregions)), function(i) {
  poly  <- ecoregions[i, ]
  ts_i  <- cd::cd_extract(catalog, poly)
  bl_i  <- cd::cd_baseline(ts_i, baseline_years = 1951:1980)
  ano_i <- cd::cd_anomaly(ts_i, bl_i)
  trn_i <- cd::cd_trend(ano_i, trend_start = c(1951, 1981))
  cmp_i <- cd::cd_compare(ts_i, window_a = 2015:2025, window_b = 1951:1980)
  list(ano = ano_i, trn = trn_i, cmp = cmp_i)
})
```

```{r per-ecoregion-reshape, include = FALSE}
# Pre-computed equivalent loaded from the bundled rds; the reshape
# below builds long-form tables and trend-line segments for the
# per-ecoregion facet plots.
results <- lapply(vignette_data$ecoregion, function(x) {
  list(ano = x$ano, trn = x$trn, cmp = x$cmp_pct)
})
# Reorder to match ecoregions (already by area_km2 desc on disk)
results <- results[ecoregions$code]

# Long-form labels: "OMM: Omineca Mountains"
er_labels <- setNames(
  paste0(ecoregions$code, ": ",
         tools::toTitleCase(tolower(ecoregions$name))),
  ecoregions$code
)

ano_all <- do.call(rbind, lapply(seq_along(results), function(i) {
  o <- results[[i]]$ano
  o$ecoregion <- ecoregions$code[i]
  o
}))
ano_all$ecoregion <- factor(ano_all$ecoregion, levels = ecoregions$code)

# Theil-Sen trend segments for facet overlay — one row per
# (ecoregion x trend_start) combination so both 75-yr and 45-yr lines
# render in each facet.
build_segments <- function(var) {
  do.call(rbind, lapply(seq_along(results), function(i) {
    sub <- results[[i]]$trn
    sub <- sub[sub$variable == var & sub$period == "annual", ]
    if (nrow(sub) == 0) return(NULL)
    yrs <- ano_all$year[ano_all$variable == var & ano_all$ecoregion == ecoregions$code[i]]
    data.frame(
      ecoregion = factor(ecoregions$code[i], levels = ecoregions$code),
      window = paste0(sub$trend_start, "–", max(yrs)),
      x_start = sub$trend_start,
      x_end = max(yrs),
      y_start = sub$intercept + sub$slope * sub$trend_start,
      y_end = sub$intercept + sub$slope * max(yrs)
    )
  }))
}
trn_segs_tmean   <- build_segments("tmean")
trn_segs_prcp    <- build_segments("prcp")
trn_segs_swe_max <- build_segments("swe_max")
trn_segs_doy_50  <- build_segments("snowmelt_doy_50")
```

```{r facet-tmean, fig.cap="Annual mean temperature anomaly relative to the 1951-1980 baseline, by ecoregion. Dashed line is the 75-year Theil-Sen trend (1951-present); solid line is the 45-year trend (1981-present). A solid line steeper than the dashed line indicates accelerating warming.", fig.height=6, echo = FALSE}
ggplot(ano_all[ano_all$variable == "tmean" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_tmean,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = expression("Mean temperature anomaly (" * degree * C * ")")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r facet-prcp, fig.cap="Annual precipitation anomaly (% of the 1951-1980 baseline) by ecoregion. Dashed line is the 75-year trend (1951-present); solid line is the 45-year trend (1981-present). The two northernmost ecoregions (BMP, NRM) show statistically significant precipitation increases over the full record; the others do not.", fig.height=6, echo = FALSE}
ggplot(ano_all[ano_all$variable == "prcp" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_prcp,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#4575b4", `FALSE` = "#d73027")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "Precipitation anomaly (% of baseline)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

### Snow per ecoregion

Two snow metrics are worth viewing per-ecoregion: peak snow water
equivalent (annual snowpack magnitude) and DOY-50 (the day-of-year by
which half the year's snowmelt has already happened — earlier values
mean an earlier freshet). The two figures below show how each varies
ecoregion to ecoregion. Pair them with the Watershed Groups Across
Ecoregions section below to read each watershed group's dominant
ecoregion's signal.

```{r facet-swe-max, fig.cap="Annual peak snow water equivalent (SWE) anomaly by ecoregion, relative to the 1951-1980 baseline. Bars are mm of water-equivalent snowpack departure from baseline. Dashed line is the 75-year Theil-Sen trend (1951-present); solid line is the 45-year trend (1981-present).", fig.height=6, echo = FALSE}
ggplot(ano_all[ano_all$variable == "swe_max" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_swe_max,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#4575b4", `FALSE` = "#d73027")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "Peak SWE anomaly (mm)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r facet-doy-50, fig.cap="Snowmelt 50% day-of-year (DOY-50) anomaly by ecoregion, relative to the 1951-1980 baseline. Negative values (red) mean the freshet midpoint shifted earlier in the year. Dashed line is the 75-year Theil-Sen trend; solid line is the 45-year trend.", fig.height=6, echo = FALSE}
ggplot(ano_all[ano_all$variable == "snowmelt_doy_50" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_doy_50,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "DOY-50 anomaly (days; negative = earlier melt)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

### Snow per watershed group

The four watershed groups that make up the AOI map onto the FWCP
reporting unit directly. Per-WSG facet plots show the same
two-metric snow signal — peak SWE and DOY-50 — broken out by
watershed group rather than by ecoregion.

```{r per-wsg-recipe, eval = FALSE}
# Same cd pipeline applied per watershed group polygon:
wsg_results <- lapply(seq_len(nrow(wsgs)), function(i) {
  poly  <- wsgs[i, ]
  ts_i  <- cd::cd_extract(catalog, poly)
  bl_i  <- cd::cd_baseline(ts_i, baseline_years = 1951:1980)
  ano_i <- cd::cd_anomaly(ts_i, bl_i)
  trn_i <- cd::cd_trend(ano_i, trend_start = c(1951, 1981))
  list(ano = ano_i, trn = trn_i)
})
```

```{r per-wsg-reshape, include = FALSE}
wsg_results <- vignette_data$wsg
wsg_codes_v <- vignette_data$wsg_codes

ano_wsg <- do.call(rbind, lapply(seq_along(wsg_results), function(i) {
  o <- wsg_results[[i]]$ano
  o$wsg <- wsg_codes_v[i]
  o
}))
ano_wsg$wsg <- factor(ano_wsg$wsg, levels = wsg_codes_v)

build_segments_wsg <- function(var) {
  do.call(rbind, lapply(seq_along(wsg_results), function(i) {
    sub <- wsg_results[[i]]$trn
    sub <- sub[sub$variable == var & sub$period == "annual", ]
    if (nrow(sub) == 0) return(NULL)
    yrs <- ano_wsg$year[ano_wsg$variable == var & ano_wsg$wsg == wsg_codes_v[i]]
    data.frame(
      wsg = factor(wsg_codes_v[i], levels = wsg_codes_v),
      window = paste0(sub$trend_start, "–", max(yrs)),
      x_start = sub$trend_start,
      x_end = max(yrs),
      y_start = sub$intercept + sub$slope * sub$trend_start,
      y_end = sub$intercept + sub$slope * max(yrs)
    )
  }))
}
trn_wsg_swe_max <- build_segments_wsg("swe_max")
trn_wsg_doy_50  <- build_segments_wsg("snowmelt_doy_50")
```

```{r facet-wsg-swe-max, fig.cap="Annual peak SWE anomaly by watershed group. Bars are mm of water-equivalent snowpack departure from the 1951-1980 baseline. Dashed line is the 75-year Theil-Sen trend; solid line is the 45-year trend.", fig.height=5, echo = FALSE}
ggplot(ano_wsg[ano_wsg$variable == "swe_max" & ano_wsg$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_wsg_swe_max,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#4575b4", `FALSE` = "#d73027")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ wsg, ncol = 2) +
  labs(x = NULL, y = "Peak SWE anomaly (mm)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r facet-wsg-doy-50, fig.cap="Snowmelt DOY-50 anomaly by watershed group. Negative values (red) mean the freshet midpoint shifted earlier in the year.", fig.height=5, echo = FALSE}
ggplot(ano_wsg[ano_wsg$variable == "snowmelt_doy_50" & ano_wsg$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_wsg_doy_50,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ wsg, ncol = 2) +
  labs(x = NULL, y = "DOY-50 anomaly (days; negative = earlier melt)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r rollup, echo = FALSE}
get_75 <- function(trn, var) {
  trn[trn$variable == var & trn$period == "annual" & trn$trend_start == 1951, ]
}
rollup <- do.call(rbind, lapply(seq_along(results), function(i) {
  res <- results[[i]]
  cmp <- res$cmp
  tmean <- get_75(res$trn, "tmean")
  tmax  <- get_75(res$trn, "tmax")
  tmin  <- get_75(res$trn, "tmin")
  prcp  <- get_75(res$trn, "prcp")
  vpd   <- get_75(res$trn, "vpd")
  data.frame(
    Ecoregion = ecoregions$code[i],
    `tmean degC/dec` = round(10 * tmean$slope, 2),
    `tmax degC/dec`  = round(10 * tmax$slope, 2),
    `tmin degC/dec`  = round(10 * tmin$slope, 2),
    `prcp mm/yr` = round(prcp$slope, 3),
    `prcp p` = round(prcp$mk_pvalue, 3),
    `vpd hPa/dec` = round(10 * vpd$slope, 3),
    `vpd p` = round(vpd$mk_pvalue, 3),
    `prcp pct change` = round(
      cmp$difference[cmp$variable == "prcp" & cmp$period == "annual"], 1
    ),
    `soil moisture pct change` = round(
      cmp$difference[cmp$variable == "soil_moisture" & cmp$period == "annual"], 1
    ),
    check.names = FALSE
  )
}))
knitr::kable(rollup, row.names = FALSE,
  caption = "Per-ecoregion roll-up over the 75-year window (1951-present): annual mean, daytime maximum, and overnight minimum temperature trends (degrees C per decade); annual precipitation trend (mm per year) and Mann-Kendall p-value; annual vapour pressure deficit trend (hPa per decade) and p-value; and recent (2015-2025) vs pre-warming (1951-1980) percent change for precipitation and soil moisture. All temperature p-values are below 0.001. A tmin slope greater than the tmax slope indicates the textbook day-night asymmetry.")
```

Ecoregion codes used in the table above: `r paste0(ecoregions$code, " — ", ecoregions$name_tc, collapse = "; ")`.

<br>

## Watershed Groups Across Ecoregions

The Kootenay Lake Region is reported and managed at the watershed-group
scale — the British Columbia Freshwater Atlas hydrological reporting
unit, and the unit the Fish & Wildlife Compensation Program funds
work in. Four watershed groups make up the AOI for this
list. They span the four ecoregions in different ways. KOTL, DUNC,
and SLOC sit almost entirely within Northern Columbia Mountains
(NCM); LARL is the most-mixed group, splitting across NCM and the
Selkirk-Bitterroot Foothills (SBF). The map below shows the
watershed groups labelled with their codes, on top of ecoregion
fills. The table that follows gives the share of each watershed
group's area falling in each ecoregion.

```{r map-wsgs, fig.cap="The four watershed groups making up the AOI (KOTL, LARL, DUNC, SLOC) on top of ecoregion fills. Northern Columbia Mountains (NCM) covers most of the area; Selkirk-Bitterroot Foothills (SBF) covers the western portion of LARL; small slivers in the southwest belong to Thompson-Okanagan Plateau (TOP) and Pacific and Cascade Ranges (PTR).", fig.height=7, echo = FALSE}
wsgs <- st_read(ctx, layer = "wsgs", quiet = TRUE)

wsg_centroids <- suppressWarnings(st_centroid(wsgs))

ggplot() +
  geom_sf(data = ecoregions, aes(fill = name_tc), color = "grey45",
          linewidth = 0.3, alpha = 0.45) +
  geom_sf(data = aoi, fill = NA, color = "#2166ac", linewidth = 0.8) +
  geom_sf(data = wsgs, fill = NA, color = "grey25", linewidth = 0.4) +
  ggrepel::geom_label_repel(
    data = wsg_centroids,
    aes(label = code, geometry = geom),
    stat = "sf_coordinates",
    size = 2.6, fill = "white", alpha = 0.85,
    label.padding = unit(0.15, "lines"),
    min.segment.length = 0
  ) +
  scale_fill_brewer(palette = "Set3", name = "Ecoregion") +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_sf(
    xlim = c(aoi_bb["xmin"] - 0.4, aoi_bb["xmax"] + 0.4),
    ylim = c(aoi_bb["ymin"] - 0.5, aoi_bb["ymax"] + 0.3)
  ) +
  labs(title = "Watershed Groups on Ecoregion Backdrop") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.key.size = unit(0.4, "cm"))
```

```{r wsg-ecoregion-table, echo = FALSE}
# Pre-computed inline by data-raw/kootenay_lake_vignette_data.R
ovr <- vignette_data$wsg_eco_overlap
ovr_wide <- reshape(
  ovr[, c("wsg_code", "ecoregion_code", "pct_of_wsg")],
  idvar = "wsg_code", timevar = "ecoregion_code", direction = "wide"
)
names(ovr_wide) <- gsub("^pct_of_wsg\\.", "", names(ovr_wide))
# Order WSG rows alphabetically; replace NAs with 0
ovr_wide <- ovr_wide[order(ovr_wide$wsg_code), ]
for (col in setdiff(names(ovr_wide), "wsg_code")) {
  ovr_wide[[col]][is.na(ovr_wide[[col]])] <- 0
}
# Add full WSG name
wsg_names <- setNames(ovr$wsg_name, ovr$wsg_code)[!duplicated(ovr$wsg_code)]
ovr_wide$Name <- wsg_names[ovr_wide$wsg_code]
ovr_wide <- ovr_wide[, c("wsg_code", "Name",
                          intersect(c("NCM","SBF","TOP","PTR"), names(ovr_wide)))]
names(ovr_wide)[1] <- "Code"
for (col in setdiff(names(ovr_wide), c("Code", "Name"))) {
  names(ovr_wide)[names(ovr_wide) == col] <- paste0(col, " %")
}

kableExtra::kable_styling(
  knitr::kable(ovr_wide, label = NA,
    caption = "Watershed groups in the Kootenay Lake Region with the percent of each group's area falling in each ecoregion. KOTL, DUNC, and SLOC are essentially within Northern Columbia Mountains; LARL is the only WSG that meaningfully splits across two ecoregions (NCM + SBF), making it the place where ecoregion-level findings can pull in two directions.",
    row.names = FALSE),
  bootstrap_options = c("striped", "hover", "condensed")
)
```

<br>

## Interpretation

Three findings emerge from the climate departures shown above.

**Warming is broad and fast.** All four ecoregions warmed between
+1.6 °C and +1.9 °C cumulative since 1951, at a rate near
+0.25 °C per decade, with Mann-Kendall p-values below 0.001 in every
ecoregion. The trend is statistically significant beyond reasonable
doubt of being random noise. Daily maximum, daily minimum, and daily
mean temperatures all tell the same story. The whole region moved
together — there is no warming hot-spot at the ecoregion scale.

**Precipitation is declining and the decline is regionally
consistent.** Annual precipitation dropped **~7%** from the
1951-1980 reference to the 2015-2025 recent decade, with a long-term
Mann-Kendall p of 0.02. Spatially the decline is consistent across
all four ecoregions of the AOI, so the pattern at the regional
average mirrors what each watershed group sees individually. All
four watershed groups in the AOI are sitting in the same drying
signal.

**The atmosphere is drying.** Vapour pressure deficit — the gap
between how much water the air could hold and how much it actually
does — is up significantly across the region, mirroring the
continental-scale drying that @ficklin_novick2017Historicprojected
documented for the United States as a whole, with the strongest
historical VPD increases concentrated in the western and southern
portions, driven by combined air-temperature increases and
relative-humidity declines. Combined
with declining precipitation, this is a "double-dipping" signal:
less water arriving as precipitation, *and* warmer air pulling more
water out of soil and vegetation through evapotranspiration. Soil
moisture is essentially flat despite the precipitation decline —
the warmer atmosphere is drinking surface moisture at a rate that
keeps the topsoil layer roughly even, but the resulting
late-summer water deficit shows up in stream baseflow rather than
in soil-moisture statistics.

**Snow is leaving *and* falling less.** Annual snowfall is
**down 15%** alongside the freshet shift — winter precipitation is
falling more often as rain instead of snow, matching the threshold
finding from Knowles et al.
[-@knowles_etal2006TrendsSnowfall] for sites with winter wet-day
minimum temperatures warmer than -5 °C. Annual snow water
equivalent (SWE) is down 23% (206 → 160 mm); summer SWE has
collapsed by 73%; the snowmelt 50% day-of-year shifted **12.6
days earlier** between the reference and recent windows. The
per-WSG facet plot above shows the freshet-timing shift clearly
in all four watershed groups; the per-ecoregion breakdown shows
the signal is not concentrated in any one ecoregion.

For the cold-water resident salmonids the Kootenay Lake region
supports — bull trout, Gerrard rainbow trout, mountain whitefish,
kokanee — these signals compound. Stream temperatures track ambient
air temperatures, so warmer air means warmer water. Cold-water fish
lose thermal habitat under that forcing more than warm-water fish do
[@eaton_scheller1996Effectsclimate], and warming summer streams
combined with altered low flows reduce salmonid reproductive success
[@mantua_etal2010Climatechange]. Three knock-on effects matter here.
**Late-summer low flows are not being relieved by precipitation** —
rain is falling, not rising. **The cold-water input that
high-elevation snowpack provides during the warmest weeks of summer
is dropping in parallel with summer SWE.** **The spring freshet — the
high-flow event that shapes channels, moves spawning gravels, and
refills off-channel rearing habitat — is arriving weeks earlier.**

Lower Columbia River reaches below Hugh Keenleyside Dam are
dam-fragmented and not anadromous, so the framing is about resident
salmonids and their habitat rather than salmon migrations.

## References
