---
title: "Climate Departure for the FWCP Peace Region"
output:
  bookdown::html_document2:
    number_sections: false
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Climate Departure for the FWCP Peace Region}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r libraries, include = FALSE}
library(terra)
library(ggplot2)
library(tidyterra)
```

The `cd` package builds on ERA5-Land hourly reanalysis (1950–present,
~9 km native grid) fetched from the DestinE Earth Data Hub, aggregated
to monthly, seasonal, and annual periods over British Columbia, and
published as Cloud-Optimized GeoTIFFs alongside a static STAC catalog
in a public S3 bucket. R consumer functions read those GeoTIFFs
directly via GDAL, crop to any area of interest, and compute baselines,
anomalies, and Mann-Kendall / Theil-Sen trend statistics.

This vignette runs the consumer pipeline on a regional administrative
boundary — the Fish & Wildlife Compensation Program (FWCP) Peace Region,
~73,000 km² in northeastern British Columbia covering Williston Reservoir
and surrounding watersheds.

## Area of Interest

The bundled area of interest is the FWCP Peace Region polygon (single
multi-polygon, EPSG:4326). Any `sf` polygon works. The watersheds
within this region all drain to the Mackenzie River — the Peace flows
east into the Slave, which flows into the Mackenzie, which empties
into the Beaufort Sea on the Arctic Ocean.

We also carry British Columbia ecoregions through the analysis.
Ecoregions partition the province into broad climate-physiography
zones based on latitude, elevation, and dominant vegetation. Climate
departure can vary across them in ways the regional average hides,
and we use them later as the sub-region for the per-ecoregion
breakdown.

```{r load-aoi}
library(cd)
library(sf)

aoi <- st_read(
  system.file("extdata", "example_aoi_fwcp_peace.gpkg", package = "cd"),
  quiet = TRUE
)

area_km2 <- as.numeric(sum(st_area(st_transform(aoi, 3005)))) / 1e6
```

```{r load-context, include = FALSE}
ctx <- system.file("extdata", "context_fwcp_peace.gpkg", package = "cd")
towns       <- st_read(ctx, layer = "towns",      quiet = TRUE)
lakes       <- st_read(ctx, layer = "lakes",      quiet = TRUE)
rivers      <- st_read(ctx, layer = "rivers",     quiet = TRUE)
streams     <- st_read(ctx, layer = "streams",    quiet = TRUE)
highways    <- st_read(ctx, layer = "highways",   quiet = TRUE)
ecoregions  <- st_read(ctx, layer = "ecoregions", quiet = TRUE)
ecoregions  <- ecoregions[ecoregions$area_km2 > 100, ]
ecoregions$name_tc <- tools::toTitleCase(tolower(ecoregions$name))
```

```{r map-location, fig.cap="FWCP Peace Region area of interest (~73,000 km^2) in northeastern British Columbia, coloured by ecoregion. Williston Reservoir dominates the basin.", fig.height=7}
aoi_bb <- st_bbox(aoi)

ggplot() +
  geom_sf(data = ecoregions, aes(fill = name_tc), color = "grey45",
          linewidth = 0.3, alpha = 0.45) +
  geom_sf(data = aoi, fill = NA, color = "#2166ac", linewidth = 0.8) +
  geom_sf(data = lakes, fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = rivers, fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = streams, color = "#74a9cf", linewidth = 0.3, alpha = 0.6) +
  geom_sf(data = highways, color = "#333333", linewidth = 0.5) +
  geom_sf(data = towns, color = "black", size = 2.5) +
  ggrepel::geom_label_repel(
    data = towns,
    aes(label = name, geometry = geom),
    stat = "sf_coordinates",
    size = 3.2, fill = "white", alpha = 0.85,
    label.padding = unit(0.2, "lines"),
    min.segment.length = 0
  ) +
  scale_fill_brewer(palette = "Set3", name = "Ecoregion") +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_sf(
    xlim = c(aoi_bb["xmin"] - 0.4, aoi_bb["xmax"] + 0.4),
    ylim = c(aoi_bb["ymin"] - 0.5, aoi_bb["ymax"] + 0.3)
  ) +
  labs(title = "FWCP Peace Region",
       subtitle = paste0(round(area_km2), " km^2 — ", nrow(ecoregions), " ecoregions")) +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.key.size = unit(0.4, "cm"))
```

## Connect to the Data Catalog

The producer pipeline in this repository fetches ERA5-Land hourly
reanalysis data, derives variables (vapour pressure deficit, relative
humidity, soil moisture), aggregates to monthly / seasonal / annual,
and writes Cloud-Optimized GeoTIFFs to S3. Alongside the GeoTIFFs we
publish a static SpatioTemporal Asset Catalog (STAC) — a small set of
JSON files that index the assets. Both the GeoTIFFs and the catalog
JSON live at
<https://stac-era5-land.s3.us-west-2.amazonaws.com/catalog.json>.

`cd_catalog()` reads that URL by default. The Cloud-Optimized GeoTIFFs
are also usable directly outside R — for example, in QGIS via the STAC
plugin, in `gdalcubes`, or with any STAC-aware client.

```{r catalog}
catalog <- cd_catalog()
kableExtra::kable_styling(
  knitr::kable(catalog, label = NA,
    caption = "Available climate variables and periods in the STAC catalog."),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "320px")
```

<br>

## Extract Climate Time Series

`cd_extract()` crops each cloud-hosted GeoTIFF to the area of interest
and computes the spatial mean per year. For an area of interest of
this size a live extraction takes a few seconds per variable. To keep
the vignette fast and reproducible we load a pre-computed result of
exactly that call.

```{r extract}
# In a fresh interactive session, you would compute this with:
#   ts <- cd_extract(catalog, aoi)
# Below we load the pre-computed equivalent so the vignette renders
# without hitting the network. ts is the time series tibble; bl_early,
# ano, trn, cmp, and cmp_pct are the downstream products.
vignette_data <- readRDS(system.file(
  "vignette-data", "peace_fwcp.rds", package = "cd"
))
ts       <- vignette_data$regional$ts
bl_early <- vignette_data$regional$bl
ano      <- vignette_data$regional$ano
trn      <- vignette_data$regional$trn
cmp      <- vignette_data$regional$cmp
cmp_pct  <- vignette_data$regional$cmp_pct

knitr::kable(head(ts, 10),
  caption = "First 10 rows of the extracted climate time series.")
```

## Trends

Anomalies are computed against a pre-warming reference period — 1951–1980,
the three decades before climate change accelerated. This is the same base
period @hansen_etal2012Perceptionclimate use to detect the emergence of
3-sigma summertime-temperature outliers globally. Saying a year is
"+1.5 °C" means it was 1.5 °C warmer than the average year between 1951
and 1980.

The trend table that follows has two rows per variable. We compute trends
from two different start years:

- **1951–present (75 years)** — the long view. Captures the full
  magnitude of warming since the pre-warming reference.
- **1981–present (45 years)** — starts at the beginning of the World
  Meteorological Organization's most recent 30-year "climate normal"
  (1981–2010) [@arguez_vose2011DefinitionStandard]. This is the
  reference period used in most published climate products, so it
  makes results easy to compare against Intergovernmental Panel on
  Climate Change reports and government climate summaries.

Comparing the two slopes is informative. If the 45-year slope is steeper
than the 75-year slope, warming has accelerated — recent decades are
heating faster than the long-term average. If the 45-year slope is
shallower, warming has slowed (though "slower" almost never means
"stopped"). When the two slopes are similar, the rate of change has been
roughly steady across the full record.

Total Change is the slope multiplied by the number of years — the
cumulative shift over the trend window.

```{r trend}
# Equivalent to:
#   bl_early <- cd_baseline(ts, baseline_years = 1951:1980)
#   ano <- cd_anomaly(ts, bl_early)
#   trn <- cd_trend(ano, trend_start = c(1951, 1981))
kableExtra::kable_styling(
  knitr::kable(cd_summary(trn), label = NA,
    caption = "Trend statistics for all variables and periods, FWCP Peace Region."),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "320px")
```

<br>

```{r plot-tmean, fig.cap="Annual mean temperature anomaly for the FWCP Peace Region relative to 1951-1980 baseline."}
cd_plot_timeseries(
  ano, variable = "tmean", period = "annual", trend = trn,
  title = "Annual Mean Temperature Anomaly — FWCP Peace Region"
)
```

```{r plot-prcp, fig.cap="Annual precipitation anomaly (% of 1951-1980 baseline) for the FWCP Peace Region."}
cd_plot_timeseries(
  ano, variable = "prcp", period = "annual", trend = trn,
  title = "Annual Precipitation Anomaly — FWCP Peace Region"
)
```

## Daytime Highs and Overnight Lows

The cd package ships daytime maximum (tmax) and overnight minimum (tmin)
temperatures alongside the daily mean. They carry distinct information.
Overnight minimums warming faster than daytime maximums — the
"day-night asymmetry" — is one of the textbook fingerprints of
greenhouse warming [@karl_etal1993NewPerspective]. Whether a
watershed or region shows that signal depends on local geography
(valley inversions, snow cover, slope-aspect mix).

For the FWCP Peace Region, **overnight minimums are warming faster
than daytime maximums** — the textbook day-night asymmetry. Daytime
maximums warmed about +0.027 °C per year since 1951 (+2.0 °C
cumulative), while overnight minimums warmed about +0.032 °C per year
(+2.4 °C cumulative). The overnight side warmed roughly 0.4 °C more
than the daytime side over the full record, narrowing the diurnal
temperature range by the same amount. The three figures below show
the tmax, tmin and diurnal-range time series that yield those
numbers.

```{r plot-tmax, fig.cap="Annual daytime maximum temperature (tmax) anomaly for the FWCP Peace Region relative to the 1951-1980 baseline."}
trn_tmax <- cd_trend(
  ano[ano$variable == "tmax" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "tmax", period = "annual", trend = trn_tmax,
  title = "Daytime Maximum (tmax) — Annual Anomaly")
```

```{r plot-tmin, fig.cap="Annual overnight minimum temperature (tmin) anomaly for the FWCP Peace Region relative to the 1951-1980 baseline."}
trn_tmin <- cd_trend(
  ano[ano$variable == "tmin" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "tmin", period = "annual", trend = trn_tmin,
  title = "Overnight Minimum (tmin) — Annual Anomaly")
```

```{r plot-dtr, fig.cap="Diurnal temperature range (daytime maximum minus overnight minimum) annual mean for the FWCP Peace Region. The downward trend indicates overnight lows are warming faster than daytime highs — the textbook day-night asymmetry shows up here."}
tmax_ts <- ts[ts$variable == "tmax" & ts$period == "annual", c("year", "value")]
tmin_ts <- ts[ts$variable == "tmin" & ts$period == "annual", c("year", "value")]
dtr <- merge(tmax_ts, tmin_ts, by = "year", suffixes = c("_max", "_min"))
dtr$dtr <- dtr$value_max - dtr$value_min

ggplot(dtr, aes(x = year, y = dtr)) +
  geom_line(color = "grey50") +
  geom_point(color = "grey30", size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "#b2182b", linewidth = 0.8) +
  labs(
    title = "Diurnal Temperature Range — FWCP Peace Region",
    x = NULL,
    y = expression("Daytime maximum minus overnight minimum (" * degree * "C)")
  ) +
  theme_minimal(base_size = 12)
```

## Snowpack

Snowpack is the hinge of BC hydrology: winter precipitation falls as
snow, accumulates on the ground, and releases as meltwater across
spring and summer. That seasonal storage is the difference between a
late-summer creek that still flows and one that doesn't. It also sets
the timing of the spring freshet — the annual flood pulse that
salmonids time their up-river migrations to. Cordillera-wide,
snowpack has been declining for decades [@mote_etal2018Dramaticdeclines;
@pederson_etal2011UnusualNature]. For four BC river basins (Fraser,
Peace, Columbia, Campbell), Najafi et al. [-@najafi_etal2017AttributionObserved]
attribute observed spring SWE decline to anthropogenic forcing using
formal detection-attribution. For the Fraser specifically, Kang et al.
[-@kang_etal2016ImpactsRapidly] document a ~10-day advance of the
spring freshet over 1949-2006, with declining summer flows during the
salmon migration window.

**For the FWCP Peace Region, the snowpack signal in our data is
unambiguous and concentrated in the warm shoulders of the year.**
Annual snow water equivalent (SWE) declined ~10% (135 → 122 mm),
but the seasonal breakdown is sharper: **summer SWE collapsed by
75%** (21.5 → 5.3 mm) and **spring snowmelt rose 37%** (212 → 290 mm)
as the freshet shifted earlier into the calendar. Total annual
snowfall barely changed (-6%) — the SWE decline is mostly about
warming removing snow earlier, not less snow falling.

A few notes on how to read the snow numbers below.

ERA5-Land represents snow on a roughly 9 km grid — each grid cell is
a single number summarizing snow averaged over about 80 km² of mixed
terrain (saddles, slopes, valleys, forest, exposed alpine, all
combined). When we compare the model against a snow station inside
that cell, two distinct errors can stack: a **scale mismatch** (a
single point measurement isn't the same thing as an 80 km² average,
especially in mountain terrain where snow accumulation varies
sharply over short distances), and a **cell-mean bias** that the
model has at the Northern Hemisphere scale (ERA5-Land overestimates
mountain SWE by 150-200% even when compared against area-averaged
satellite estimates [@kouki_etal2023Evaluationsnow], traced to a
simplified snow layer in the underlying atmospheric model).

Our QA cross-check at four British Columbia automated snow stations
inside the FWCP Peace Region (1985-2025, 95 paired station-years)
sees both kinds of error stacked and cannot fully separate them. At
Pine Pass — a Coast Mountain saddle that catches orographic precip
from systems coming inland from the Pacific — the model is 61% too
low at the station, likely mostly scale mismatch: the station sits
at a high-snow microsite within an averaged cell. At Aiken Lake — a
drier interior valley basin — the model is 54% too high, likely a
mix of scale mismatch (averaging snowier surrounding terrain into
the same cell) and the Kouki-style cell-mean bias for
interior-continental cells.

What does survive both kinds of error: **the gap between the model
and the stations is the same size in 2020 as it was in 1990.** The
bias is stable over time at every site (regression of
model-minus-station on year is non-significant, p > 0.2). That means
the *changes over time* shown below — peak snowpack dropping,
freshet shifting earlier — are real even though the absolute mm
numbers shouldn't be quoted as ground truth. For regional aggregates
(the numbers in this section are spatial means over hundreds of
cells), random point-vs-cell mismatches partially cancel, and stable
cell-mean biases preserve the trend signal.

The trend tests below use raw Mann-Kendall plus Theil-Sen — no
pre-whitening — which is the right call for our 76-year series with
strong climate trends per Yue and Wang
[-@yue_wang2002Applicabilityprewhitening] (pre-whitening
underestimates slope when a real trend exists).

### Seasonal snowpack curve

The four monthly-native snow variables — SWE, snowfall, snowmelt, and
snow cover — show *when* snow accumulates and melts. Aggregated to
the four standard meteorological seasons — winter (December–February),
spring (March–May), summer (June–August), and fall
(September–November) — the table below compares the recent decade
(2015-2025) against the pre-warming reference (1951-1980) directly.
The headline numbers above (summer SWE collapse, spring snowmelt rise)
are in the **summer** and **spring** rows.

```{r snow-seasonal-table}
snow_monthly <- c("swe", "snowfall", "snowmelt", "snow_cover")
season_order <- c("winter", "spring", "summer", "fall")

snow_seasonal <- cmp_pct[cmp_pct$variable %in% snow_monthly &
                          cmp_pct$period %in% season_order, ]
snow_seasonal$period <- factor(snow_seasonal$period, levels = season_order)
snow_seasonal$variable <- factor(snow_seasonal$variable, levels = snow_monthly,
                                  labels = c("SWE (mm)", "Snowfall (mm)",
                                             "Snowmelt (mm)", "Snow cover (%)"))
snow_seasonal <- snow_seasonal[order(snow_seasonal$variable, snow_seasonal$period), ]
snow_seasonal$mean_a <- round(snow_seasonal$mean_a, 1)
snow_seasonal$mean_b <- round(snow_seasonal$mean_b, 1)
snow_seasonal$difference <- round(snow_seasonal$difference, 1)
snow_seasonal <- snow_seasonal[, c("variable", "period", "mean_b",
                                    "mean_a", "difference")]
names(snow_seasonal) <- c("Variable", "Season",
                          "Pre-warming (1951–1980)",
                          "Recent (2015–2025)", "Δ %")

kableExtra::kable_styling(
  knitr::kable(snow_seasonal, label = NA,
    caption = "Seasonal snowpack: recent decade compared to pre-warming reference for the FWCP Peace Region. Summer SWE collapse (-75%) and spring snowmelt rise (+37%) are the headline signals.",
    row.names = FALSE),
  bootstrap_options = c("striped", "hover", "condensed")
)
```

<br>

### Annual climate-departure signals

Four derived annual scalars capture the climate-departure signals the
literature treats as headline metrics for snow hydrology: peak
snowpack [@mote_etal2018Dramaticdeclines; @mote_etal2005DECLININGMOUNTAIN],
the date of melt midpoint
[@stewart_etal2005ChangesEarlier; @cayan_etal2001ChangesOnset],
freshet flashiness, and the
fraction of precipitation falling as snow [@knowles_etal2006TrendsSnowfall].
Two notes on our specific implementation. We use the actual annual
maximum of daily SWE rather than the April-1 SWE canon
[@pederson_etal2011UnusualNature] — equivalent in effect for BC
pixels (peak is at or near April 1) but date-insensitive. And our
freshet-flashiness metric (annual maximum of 7-day rolling daily
snowmelt) is upstream of the streamflow-based flashiness measures in
the literature [@stewart_etal2005ChangesEarlier;
@kang_etal2016ImpactsRapidly] — diagnostic of snowpack-side intensity
before routing through soil and channel storage.

```{r snow-swe-max, fig.cap="Annual peak snow water equivalent (swe_max) for the FWCP Peace Region. ERA5-Land mm SWE (regional spatial mean)."}
trn_swe_max <- cd_trend(
  ano[ano$variable == "swe_max" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "swe_max", period = "annual",
                   trend = trn_swe_max,
                   title = "Annual peak SWE — Anomaly")
```

```{r snow-doy-50, fig.cap="Day of year when half the annual snowmelt has accumulated (snowmelt_doy_50). Earlier dates indicate an earlier freshet centroid."}
trn_doy <- cd_trend(
  ano[ano$variable == "snowmelt_doy_50" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "snowmelt_doy_50", period = "annual",
                   trend = trn_doy,
                   title = "Snowmelt 50% DOY — Anomaly")
```

```{r snow-rate-peak, fig.cap="Annual maximum of 7-day rolling daily snowmelt (snowmelt_rate_peak). Higher values indicate more concentrated freshet pulses."}
trn_rate <- cd_trend(
  ano[ano$variable == "snowmelt_rate_peak" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "snowmelt_rate_peak", period = "annual",
                   trend = trn_rate,
                   title = "Peak weekly melt rate — Anomaly")
```

```{r snow-fraction, fig.cap="Annual snowfall fraction (snowfall_fraction): the percent of annual precipitation that fell as snow. Anomaly is in percentage points."}
trn_frac <- cd_trend(
  ano[ano$variable == "snowfall_fraction" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "snowfall_fraction", period = "annual",
                   trend = trn_frac,
                   title = "Snowfall fraction — Anomaly")
```

### What this means for the FWCP Peace Region

Three findings carry the snowpack story for the Peace.

**Snow is leaving the region earlier each year, not falling
less.** Total annual snowfall is roughly stable (-6%); the snowpack
decline is dominated by *earlier* melt, not by *less* snow. The
seasonal data make this concrete: spring snowmelt is up 37% while
summer snowmelt is down 61% — same total melt, redistributed earlier
into the year. This matches the broader Pacific-Northwest signal
documented since Mote et al. [-@mote_etal2005DECLININGMOUNTAIN].

**The freshet is shifting into spring.** Spring snowmelt up 37% is
direct evidence of an earlier freshet centroid — consistent in
direction and rough magnitude with Stewart et al.'s
[-@stewart_etal2005ChangesEarlier] 1-4 week earlier streamflow
timing across western North America, and with the 10-day Fraser
freshet advance documented by Kang et al.
[-@kang_etal2016ImpactsRapidly] for a basin whose southern boundary
is just south of the FWCP Peace Region.

**Summer is becoming snow-free.** Summer SWE has collapsed by 75%
and summer snowmelt is down 61%. The high-elevation snowpack that
historically lingered into summer no longer does in the recent
decade. For aquatic ecosystems downstream, this is a loss of late-
season cold-water input to streams during the warmest part of the
year.

## Recent vs Pre-warming

The table below compares two windows directly — the recent decade
(2015–2025) against the pre-warming reference (1951–1980). Δ
absolute is the difference of the two means in the variable's native
units. Δ % is shown only for variables where it is meaningful
(precipitation, soil moisture, vapour pressure deficit, relative
humidity). The trend p column is the Mann-Kendall p-value of the
75-year (1951–present) trend on the same variable and period; it
tests for a steady year-on-year ramp, which is a related but distinct
question from "do these two windows differ".

The recent decade was 1.7 to 2.4 °C warmer than the pre-warming
reference for annual mean, daytime maximum, and overnight minimum,
with Mann-Kendall trend p-values below 0.001. Vapour pressure deficit
is up significantly. Annual precipitation was about 3 to 4 % higher
in the recent decade; the long-term trend test does not confirm a
steady year-on-year ramp (p ≈ 0.20). Soil moisture and relative
humidity show no meaningful change.

```{r compare}
trn_p <- trn[trn$trend_start == 1951, c("variable", "period", "mk_pvalue")]
names(trn_p)[3] <- "trend_p"
trn_p$trend_p <- round(trn_p$trend_p, 3)

no_pct_vars <- c(
  "tmean", "tmax", "tmin",
  # snow_cover and snowfall_fraction are already in % (pct_point_diff
  # anomaly type) — pct-of-baseline mixes units. snowmelt_doy_50 is a
  # day-of-year — pct-change of a date is meaningless.
  "snow_cover", "snowfall_fraction", "snowmelt_doy_50"
)
cmp_combined <- cmp |>
  dplyr::mutate(
    pct_change = ifelse(
      variable %in% no_pct_vars,
      NA_real_,
      round(100 * (mean_a - mean_b) / mean_b, 1)
    ),
    mean_a   = round(mean_a, 2),
    mean_b   = round(mean_b, 2),
    abs_diff = round(difference, 2)
  ) |>
  dplyr::select(variable, period, mean_a, mean_b, abs_diff, pct_change) |>
  merge(trn_p, by = c("variable", "period"), all.x = TRUE)

names(cmp_combined) <- c(
  "Variable", "Period",
  "Recent (2015–2025)", "Pre-warming (1951–1980)",
  "Δ absolute", "Δ %", "Trend p (75-yr)"
)

kableExtra::kable_styling(
  knitr::kable(cmp_combined, label = NA,
    caption = "Recent decade (2015-2025 mean) compared to pre-warming reference (1951-1980 mean) for the FWCP Peace Region.",
    row.names = FALSE),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "360px")
```

<br>

## Spatial Pattern

The zonal mean reduces a region this size to a single number; the spatial
pattern carries the rest of the story.

Warming is not spatially uniform. Total departures range from about
+1.5 °C in the southeast to over +2.1 °C in the west and centre. The
dominant gradient runs east-to-west — the western half of the region
averages roughly 0.2 °C more warming than the eastern half — with a
smaller south-to-north component. The west-of-Rockies pattern is
consistent with windward-slope amplification along the Continental
Divide.

```{r spatial-tmean, fig.cap="Spatial pattern of annual mean temperature departure across the FWCP Peace Region (2015-2025 mean minus 1951-1980 mean, degrees C).", fig.height=7}
# Pre-computed by data-raw/peace_fwcp_vignette_data.R. Live equivalent:
#   r_tmean <- cd_crop(catalog$href[catalog$variable == "tmean"
#                                   & catalog$period == "annual"], aoi)
#   years <- as.integer(names(r_tmean))
#   departure <- mean(r_tmean[[which(years >= 2015 & years <= 2025)]]) -
#                mean(r_tmean[[which(years >= 1951 & years <= 1980)]])
#   departure <- terra::mask(departure, aoi)
departure <- terra::rast(system.file(
  "vignette-data", "peace_fwcp_departure_tmean.tif", package = "cd"
))
names(departure) <- "Temperature departure"

ggplot() +
  geom_spatraster(data = departure) +
  geom_sf(data = ecoregions, fill = NA, color = "grey25",
          linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = aoi, fill = NA, color = "black", linewidth = 0.6) +
  geom_sf(data = lakes, fill = NA, color = "grey40", linewidth = 0.2) +
  geom_sf(data = highways, color = "#333333", linewidth = 0.4) +
  geom_sf(data = towns, color = "black", size = 2) +
  ggrepel::geom_label_repel(
    data = towns, aes(label = name, geometry = geom),
    stat = "sf_coordinates", size = 3, fill = "white", alpha = 0.8
  ) +
  scale_fill_distiller(
    palette = "RdBu", direction = -1,
    name = expression(Delta * degree * C)
  ) +
  coord_sf(
    xlim = c(aoi_bb["xmin"], aoi_bb["xmax"]),
    ylim = c(aoi_bb["ymin"], aoi_bb["ymax"])
  ) +
  labs(title = "Temperature Departure (2015-2025 vs 1951-1980)") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank())
```

## Per-Ecoregion Variation

The regional zonal mean averages over five ecoregions with different
elevations and exposures. To check whether the regional story holds
within each ecoregion — and where it does not — we run the same
pipeline on each ecoregion polygon individually.

All five ecoregions warmed at roughly the same rate — about +0.3 °C
per decade since 1951, with cumulative changes within 0.2 °C of each
other. Precipitation is the variable that diverges: the two
northernmost ecoregions (Boreal Mountains and Plateaus, Northern
Canadian Rocky Mountains) show statistically significant precipitation
increases, while the southern and central ecoregions do not. Vapour
pressure deficit is up significantly in all five.

```{r per-ecoregion-compute, include = FALSE}
# Pre-computed by data-raw/peace_fwcp_vignette_data.R. Live equivalent
# would loop cd_extract / cd_baseline / cd_anomaly / cd_trend / cd_compare
# over each ecoregion polygon.
results <- lapply(vignette_data$ecoregion, function(x) {
  list(ano = x$ano, trn = x$trn, cmp = x$cmp_pct)
})
# Reorder to match ecoregions (already by area_km2 desc on disk)
results <- results[ecoregions$code]

# Long-form labels: "OMM: Omineca Mountains"
er_labels <- setNames(
  paste0(ecoregions$code, ": ",
         tools::toTitleCase(tolower(ecoregions$name))),
  ecoregions$code
)

ano_all <- do.call(rbind, lapply(seq_along(results), function(i) {
  o <- results[[i]]$ano
  o$ecoregion <- ecoregions$code[i]
  o
}))
ano_all$ecoregion <- factor(ano_all$ecoregion, levels = ecoregions$code)

# Theil-Sen trend segments for facet overlay — one row per
# (ecoregion x trend_start) combination so both 75-yr and 45-yr lines
# render in each facet.
build_segments <- function(var) {
  do.call(rbind, lapply(seq_along(results), function(i) {
    sub <- results[[i]]$trn
    sub <- sub[sub$variable == var & sub$period == "annual", ]
    if (nrow(sub) == 0) return(NULL)
    yrs <- ano_all$year[ano_all$variable == var & ano_all$ecoregion == ecoregions$code[i]]
    data.frame(
      ecoregion = factor(ecoregions$code[i], levels = ecoregions$code),
      window = paste0(sub$trend_start, "–", max(yrs)),
      x_start = sub$trend_start,
      x_end = max(yrs),
      y_start = sub$intercept + sub$slope * sub$trend_start,
      y_end = sub$intercept + sub$slope * max(yrs)
    )
  }))
}
trn_segs_tmean   <- build_segments("tmean")
trn_segs_prcp    <- build_segments("prcp")
trn_segs_swe_max <- build_segments("swe_max")
trn_segs_doy_50  <- build_segments("snowmelt_doy_50")
```

```{r facet-tmean, fig.cap="Annual mean temperature anomaly relative to the 1951-1980 baseline, by ecoregion. Dashed line is the 75-year Theil-Sen trend (1951-present); solid line is the 45-year trend (1981-present). A solid line steeper than the dashed line indicates accelerating warming.", fig.height=6}
ggplot(ano_all[ano_all$variable == "tmean" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_tmean,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = expression("Mean temperature anomaly (" * degree * C * ")")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r facet-prcp, fig.cap="Annual precipitation anomaly (% of the 1951-1980 baseline) by ecoregion. Dashed line is the 75-year trend (1951-present); solid line is the 45-year trend (1981-present). The two northernmost ecoregions (BMP, NRM) show statistically significant precipitation increases over the full record; the others do not.", fig.height=6}
ggplot(ano_all[ano_all$variable == "prcp" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_prcp,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#4575b4", `FALSE` = "#d73027")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "Precipitation anomaly (% of baseline)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

### Snow per ecoregion

Two snow metrics are worth viewing per-ecoregion: peak snow water
equivalent (annual snowpack magnitude) and DOY-50 (the day-of-year by
which half the year's snowmelt has already happened — earlier values
mean an earlier freshet). The two figures below show how each varies
ecoregion to ecoregion. Pair them with the Watershed Groups Across
Ecoregions section below to read each watershed group's dominant
ecoregion's signal.

```{r facet-swe-max, fig.cap="Annual peak snow water equivalent (SWE) anomaly by ecoregion, relative to the 1951-1980 baseline. Bars are mm of water-equivalent snowpack departure from baseline. Dashed line is the 75-year Theil-Sen trend (1951-present); solid line is the 45-year trend (1981-present).", fig.height=6}
ggplot(ano_all[ano_all$variable == "swe_max" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_swe_max,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#4575b4", `FALSE` = "#d73027")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "Peak SWE anomaly (mm)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r facet-doy-50, fig.cap="Snowmelt 50% day-of-year (DOY-50) anomaly by ecoregion, relative to the 1951-1980 baseline. Negative values (red) mean the freshet midpoint shifted earlier in the year. Dashed line is the 75-year Theil-Sen trend; solid line is the 45-year trend.", fig.height=6}
ggplot(ano_all[ano_all$variable == "snowmelt_doy_50" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_doy_50,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "DOY-50 anomaly (days; negative = earlier melt)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r rollup}
get_75 <- function(trn, var) {
  trn[trn$variable == var & trn$period == "annual" & trn$trend_start == 1951, ]
}
rollup <- do.call(rbind, lapply(seq_along(results), function(i) {
  res <- results[[i]]
  cmp <- res$cmp
  tmean <- get_75(res$trn, "tmean")
  tmax  <- get_75(res$trn, "tmax")
  tmin  <- get_75(res$trn, "tmin")
  prcp  <- get_75(res$trn, "prcp")
  vpd   <- get_75(res$trn, "vpd")
  data.frame(
    Ecoregion = ecoregions$code[i],
    `tmean degC/dec` = round(10 * tmean$slope, 2),
    `tmax degC/dec`  = round(10 * tmax$slope, 2),
    `tmin degC/dec`  = round(10 * tmin$slope, 2),
    `prcp mm/yr` = round(prcp$slope, 3),
    `prcp p` = round(prcp$mk_pvalue, 3),
    `vpd hPa/dec` = round(10 * vpd$slope, 3),
    `vpd p` = round(vpd$mk_pvalue, 3),
    `prcp pct change` = round(
      cmp$difference[cmp$variable == "prcp" & cmp$period == "annual"], 1
    ),
    `soil moisture pct change` = round(
      cmp$difference[cmp$variable == "soil_moisture" & cmp$period == "annual"], 1
    ),
    check.names = FALSE
  )
}))
knitr::kable(rollup, row.names = FALSE,
  caption = "Per-ecoregion roll-up over the 75-year window (1951-present): annual mean, daytime maximum, and overnight minimum temperature trends (degrees C per decade); annual precipitation trend (mm per year) and Mann-Kendall p-value; annual vapour pressure deficit trend (hPa per decade) and p-value; and recent (2015-2025) vs pre-warming (1951-1980) percent change for precipitation and soil moisture. All temperature p-values are below 0.001. A tmin slope greater than the tmax slope indicates the textbook day-night asymmetry.")
```

Ecoregion codes used in the table above: `r paste0(ecoregions$code, " — ", ecoregions$name_tc, collapse = "; ")`.

<br>

## Watershed Groups Across Ecoregions

The FWCP Peace Region is reported and managed at the watershed-group
scale — the British Columbia Freshwater Atlas hydrological reporting
unit, and the unit the Fish & Wildlife Compensation Program funds
work in. Sixteen watershed groups make up the canonical FWCP Peace
list. They span the five ecoregions in different ways: some sit
entirely within one, others split across two or three. The map below
shows the watershed groups labelled with their codes, on top of the
ecoregion fills. The table that follows gives, for each watershed
group, the share of its area falling in each ecoregion.

The Upper Peace River watershed group is also intersected by the FWCP
boundary, but only at its upstream end — its main drainage runs east
beyond the region, so it is excluded here to keep the reporting unit
clean.

```{r map-wsgs, fig.cap="Watershed groups intersecting the FWCP Peace Region (full extent, including the parts spilling outside the FWCP boundary), labelled with their codes, on top of ecoregion fills. The Fraser Basin, Central Canadian Rockies, and Omineca Mountains ecoregions occupy the southern and central portions; Boreal Mountains and Plateaus and Northern Canadian Rocky Mountains sit in the north.", fig.height=7}
wsgs <- st_read(ctx, layer = "wsgs", quiet = TRUE)

wsg_centroids <- suppressWarnings(st_centroid(wsgs))

ggplot() +
  geom_sf(data = ecoregions, aes(fill = name_tc), color = "grey45",
          linewidth = 0.3, alpha = 0.45) +
  geom_sf(data = aoi, fill = NA, color = "#2166ac", linewidth = 0.8) +
  geom_sf(data = wsgs, fill = NA, color = "grey25", linewidth = 0.4) +
  ggrepel::geom_label_repel(
    data = wsg_centroids,
    aes(label = code, geometry = geom),
    stat = "sf_coordinates",
    size = 2.6, fill = "white", alpha = 0.85,
    label.padding = unit(0.15, "lines"),
    min.segment.length = 0
  ) +
  scale_fill_brewer(palette = "Set3", name = "Ecoregion") +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_sf(
    xlim = c(aoi_bb["xmin"] - 0.4, aoi_bb["xmax"] + 0.4),
    ylim = c(aoi_bb["ymin"] - 0.5, aoi_bb["ymax"] + 0.3)
  ) +
  labs(title = "Watershed Groups on Ecoregion Backdrop") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.key.size = unit(0.4, "cm"))
```

```{r wsg-ecoregion-table}
wsg_xref <- read.csv(
  system.file("extdata", "peace_wsg_ecoregion_commentary.csv", package = "cd"),
  check.names = FALSE
)
wsg_xref$commentary <- NULL
names(wsg_xref) <- c("Code", "Name", "km² in FWCP",
                     "FAB %", "CRM %", "OMM %", "BMP %", "NRM %")
kableExtra::kable_styling(
  knitr::kable(wsg_xref, label = NA,
    caption = "Watershed groups in the FWCP Peace Region with the percent of each group's area falling in each of the five ecoregions. Rows where one ecoregion is at or near 100 percent indicate watershed groups contained within a single climatic-physiographic zone; rows split across two or more ecoregions indicate watershed groups whose climate departure can pull from more than one regional pattern. Codes match those in earlier figures. Source: data-raw/peace_wsg_ecoregion_commentary.R.",
    row.names = FALSE),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "420px")
```

<br>

## Interpretation

Three findings emerge from the climate departures shown above.

**Warming is broad, fast, and uniform.** All five ecoregions warmed
between +1.7 °C and +1.9 °C cumulative since 1951, at a rate near
+0.3 °C per decade, with Mann-Kendall p-values below 0.001 in every
ecoregion. The trend is statistically significant beyond reasonable
doubt of being random noise. Daily maximum, daily minimum, and daily
mean temperatures all tell the same story. There is no thermal hot
spot. The whole region has moved together. A subtle gradient does
emerge — the northern, higher-elevation ecoregions (Omineca Mountains,
Boreal Mountains and Plateaus) warmed about 0.2 °C more than the
southern Fraser Basin, consistent with the elevation-dependent
warming signal documented at mid-latitude mountain sites
[@pepin_etal2015Elevationdependentwarming], though the regional
evidence base remains heterogeneous and not every mountain region
shows the same pattern [@rangwala_miller2012Climatechange]. The
gradient is small relative to the regional signal.

**Precipitation is increasing significantly only in the two
northernmost ecoregions.** Across the FWCP Peace as a whole, the
long-term trend in annual precipitation is not statistically
significant. Broken out by ecoregion, Boreal Mountains and Plateaus
and Northern Canadian Rocky Mountains — both in the northern half of
the region — show real increases in annual precipitation
(p = 0.02 and 0.01). The southern and central ecoregions show no
significant change. This is a north-versus-south contrast that the
regional average hides, and the place where breaking the analysis
down by ecoregion was worth doing.

Mapping that contrast onto the watershed-group reporting unit: five
watershed groups sit in or are dominated by the wetter ecoregions —
Toodoggone, Firesteel and Ingenika sit almost entirely within Boreal
Mountains and Plateaus, and Finlay and Fox are mostly in Northern
Canadian Rocky Mountains. For these, the small upward precipitation
trend is statistically defensible. The remaining 13 watershed groups
sit elsewhere — Fraser Basin, Central Canadian Rockies, Omineca
Mountains — where annual precipitation has not changed in a way
distinguishable from natural variability.

**The atmosphere is drying despite, in places, more precipitation.**
Vapour pressure deficit — the gap between how much water the air
could hold and how much it actually does — rose significantly in
every ecoregion (p < 0.005 across all five), mirroring the
continental-scale drying that @ficklin_novick2017Historicprojected
documented for the United States as a whole, driven by combined
air-temperature increases and relative-humidity changes. Warmer
air holds more water before saturating, and that pulls moisture
out of soil and vegetation through evaporation and transpiration.
Soil moisture is essentially flat in every ecoregion — even where
precipitation increased — because the warmer atmosphere is drinking
the extra water back. This is the headline ecological finding for
the region: water inputs may be rising in places, but soil and
vegetation are not seeing more available moisture.

**Snow is leaving the region earlier, not falling less.** Annual
snowfall across the FWCP Peace is essentially unchanged (-6%); the
snowpack story is about *timing*, not *quantity*. Spring snowmelt
rose 37% (212 → 290 mm) and summer snowmelt fell 61% (165 → 64 mm)
— same total annual melt, redistributed earlier in the calendar.
Annual peak snow water equivalent (SWE) is down about 10%
(135 → 122 mm), and most strikingly, **summer SWE has collapsed by
75%**: the high-elevation snowpack that historically lingered into
June and July no longer does. This is the direct mechanism behind
the freshet shift described in the snow-trend literature: warmer
winters and springs do not stop snow from falling — they shorten
how long it stays on the ground.

The freshet-timing signal is *broadly uniform across the region* —
DOY-50 (day-of-year by which half the year's snowmelt has happened)
is shifting earlier at roughly 1 day per decade in every one of the
five ecoregions, with Mann-Kendall p-values below 0.01 in each
(median p ≈ 0.005). All sixteen FWCP Peace watershed groups,
regardless of which ecoregion they sit in, are seeing the same
order-of-magnitude earlier melt. The peak-SWE signal is noisier —
small recent-decade declines show up in every ecoregion, but the
year-to-year variability is large enough that the long-term Theil-Sen
slopes do not reach statistical significance at the ecoregion scale.
The 75% summer-SWE collapse is detectable in the regional aggregate
because spatial averaging suppresses that variability.

For the cold-water resident salmonids the FWCP Peace supports —
bull trout, Arctic grayling, mountain whitefish, rainbow trout,
kokanee — these signals compound. Stream temperatures are likely
rising in step with warmer ambient air temperatures, with the
combined effects of warming summer stream temperatures and
altered low flows likely reducing thermally-suitable habitat for
cold-water species [@mantua_etal2010Climatechange;
@eaton_scheller1996Effectsclimate]. The evapotranspiration
imbalance means low-flow conditions in late summer are not being
relieved by the precipitation increase that did occur; and the
cold-water input that high-elevation snowpack provides to streams
during the warmest, most thermally stressful weeks of summer is
dropping in parallel with summer SWE. The spring freshet — the
dominant high-flow event that shapes channel morphology,
mobilizes spawning gravels, and refills off-channel rearing
habitat — is shifting weeks earlier; the neighbouring Fraser
Basin documents the same kind of freshet advance
[@kang_etal2016ImpactsRapidly] at comparable magnitude.

## References
