---
title: "Climate Departure for the FWCP Peace Region"
output:
  bookdown::html_document2:
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Climate Departure for the FWCP Peace Region}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r libraries, include = FALSE}
library(terra)
library(ggplot2)
library(tidyterra)
```

The `cd` package builds on ERA5-Land hourly reanalysis (1950–present,
~9 km native grid) fetched from the DestinE Earth Data Hub, aggregated
to monthly, seasonal, and annual periods over British Columbia, and
published as Cloud-Optimized GeoTIFFs alongside a static STAC catalog
in a public S3 bucket. R consumer functions read those GeoTIFFs
directly via GDAL, crop to any area of interest, and compute baselines,
anomalies, and Mann-Kendall / Theil-Sen trend statistics.

This vignette runs the consumer pipeline on a regional administrative
boundary — the Fish & Wildlife Compensation Program (FWCP) Peace Region,
~73,000 km² in northeastern British Columbia covering Williston Reservoir
and surrounding watersheds.

For build speed and CI robustness the heavy zonal extracts and the
spatial-pattern raster are pre-computed by
`data-raw/peace_fwcp_vignette_data.R` and shipped under
`inst/vignette-data/`. The vignette below loads those results rather
than re-fetching them from S3 on every render. Re-run the script when
the underlying STAC catalog updates.

## Area of Interest

The bundled AOI is the FWCP Peace Region polygon (single multi-polygon,
EPSG:4326). Any `sf` polygon works. The watersheds within this region
all drain to the Mackenzie River — the Peace flows east into the Slave,
which flows into the Mackenzie, which empties into the Beaufort Sea on
the Arctic Ocean. This is the Arctic-draining side of British Columbia,
distinct from the Pacific-draining watersheds south and west of the
Continental Divide.

We also carry BC ecoregions through the analysis. Ecoregions partition
the province into broad climate-physiography zones based on latitude,
elevation, and dominant vegetation. Climate departure can vary across
them in ways the regional average hides, and we use them later as the
sub-region for the per-ecoregion breakdown.

```{r load-aoi}
library(cd)
library(sf)

aoi <- st_read(
  system.file("extdata", "example_aoi_fwcp_peace.gpkg", package = "cd"),
  quiet = TRUE
)

area_km2 <- as.numeric(sum(st_area(st_transform(aoi, 3005)))) / 1e6
```

```{r load-context, include = FALSE}
ctx <- system.file("extdata", "context_fwcp_peace.gpkg", package = "cd")
towns       <- st_read(ctx, layer = "towns",      quiet = TRUE)
lakes       <- st_read(ctx, layer = "lakes",      quiet = TRUE)
rivers      <- st_read(ctx, layer = "rivers",     quiet = TRUE)
streams     <- st_read(ctx, layer = "streams",    quiet = TRUE)
highways    <- st_read(ctx, layer = "highways",   quiet = TRUE)
ecoregions  <- st_read(ctx, layer = "ecoregions", quiet = TRUE)
ecoregions  <- ecoregions[ecoregions$area_km2 > 100, ]
ecoregions$name_tc <- tools::toTitleCase(tolower(ecoregions$name))
```

```{r map-location, fig.cap="FWCP Peace Region area of interest (~73,000 km^2) in northeastern British Columbia, coloured by BC ecoregion. Williston Reservoir dominates the basin.", fig.height=7}
aoi_bb <- st_bbox(aoi)

ggplot() +
  geom_sf(data = ecoregions, aes(fill = name_tc), color = "grey45",
          linewidth = 0.3, alpha = 0.45) +
  geom_sf(data = aoi, fill = NA, color = "#2166ac", linewidth = 0.8) +
  geom_sf(data = lakes, fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = rivers, fill = "#a6bddb", color = "#74a9cf", linewidth = 0.2) +
  geom_sf(data = streams, color = "#74a9cf", linewidth = 0.3, alpha = 0.6) +
  geom_sf(data = highways, color = "#333333", linewidth = 0.5) +
  geom_sf(data = towns, color = "black", size = 2.5) +
  ggrepel::geom_label_repel(
    data = towns,
    aes(label = name, geometry = geom),
    stat = "sf_coordinates",
    size = 3.2, fill = "white", alpha = 0.85,
    label.padding = unit(0.2, "lines"),
    min.segment.length = 0
  ) +
  scale_fill_brewer(palette = "Set3", name = "Ecoregion") +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_sf(
    xlim = c(aoi_bb["xmin"] - 0.4, aoi_bb["xmax"] + 0.4),
    ylim = c(aoi_bb["ymin"] - 0.5, aoi_bb["ymax"] + 0.3)
  ) +
  labs(title = "FWCP Peace Region",
       subtitle = paste0(round(area_km2), " km^2 — ", nrow(ecoregions), " ecoregions")) +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.key.size = unit(0.4, "cm"))
```

## Connect to the Data Catalog

The producer pipeline in this repository fetches ERA5-Land hourly
reanalysis data, derives variables (vapour pressure deficit, relative
humidity, soil moisture), aggregates to monthly / seasonal / annual,
and writes Cloud-Optimized GeoTIFFs to S3. Alongside the GeoTIFFs we
publish a static SpatioTemporal Asset Catalog (STAC) — a small set of
JSON files that index the assets. Both the GeoTIFFs and the catalog
JSON live at
<https://stac-era5-land.s3.us-west-2.amazonaws.com/catalog.json>.

`cd_catalog()` reads that URL by default. The Cloud-Optimized GeoTIFFs
are also usable directly outside R — for example, in QGIS via the STAC
plugin, in `gdalcubes`, or with any STAC-aware client.

```{r catalog, tab.cap = "Available climate variables and periods in the STAC catalog."}
catalog <- cd_catalog()
kableExtra::kable_styling(
  knitr::kable(catalog, caption = NA),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "320px")
```

<br>

```{r load-vignette-data, include = FALSE}
# Pre-computed by data-raw/peace_fwcp_vignette_data.R — see note above.
vignette_data <- readRDS(system.file(
  "vignette-data", "peace_fwcp.rds", package = "cd"
))
ts       <- vignette_data$regional$ts
bl_early <- vignette_data$regional$bl
ano      <- vignette_data$regional$ano
trn      <- vignette_data$regional$trn
cmp      <- vignette_data$regional$cmp
cmp_pct  <- vignette_data$regional$cmp_pct
```

## Extract Climate Time Series

`cd_extract()` crops each cloud-hosted COG to the AOI and computes the
spatial mean per year. For an AOI of this size a live extraction takes
a few seconds per variable; here we load the pre-computed result.

```{r extract}
# Equivalent to: ts <- cd_extract(catalog, aoi)
knitr::kable(head(ts, 10),
  caption = "First 10 rows of the extracted climate time series.")
```

## Trends

Anomalies are computed against a pre-warming reference period — 1951–1980,
the three decades before climate change accelerated. Saying a year is
"+1.5 °C" means it was 1.5 °C warmer than the average year between 1951
and 1980.

The trend table that follows has two rows per variable. We compute trends
from two different start years:

- **1951–present (75 years)** — the long view. Captures the full
  magnitude of warming since the pre-warming reference.
- **1981–present (45 years)** — starts at the beginning of the World
  Meteorological Organization's most recent 30-year "climate normal"
  (1981–2010). This is the reference period used in most published
  climate products, so it makes results easy to compare against
  Intergovernmental Panel on Climate Change reports and government
  climate summaries.

Comparing the two slopes is informative. If the 45-year slope is steeper
than the 75-year slope, warming has accelerated — recent decades are
heating faster than the long-term average. If the 45-year slope is
shallower, warming has slowed (though "slower" almost never means
"stopped"). When the two slopes are similar, the rate of change has been
roughly steady across the full record.

Total Change is the slope multiplied by the number of years — the
cumulative shift over the trend window.

```{r trend, tab.cap = "Trend statistics for all variables and periods, FWCP Peace Region."}
# Equivalent to:
#   bl_early <- cd_baseline(ts, baseline_years = 1951:1980)
#   ano <- cd_anomaly(ts, bl_early)
#   trn <- cd_trend(ano, trend_start = c(1951, 1981))
kableExtra::kable_styling(
  knitr::kable(cd_summary(trn), caption = NA),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "320px")
```

<br>

```{r plot-tmean, fig.cap="Annual mean temperature anomaly for the FWCP Peace Region relative to 1951-1980 baseline."}
cd_plot_timeseries(
  ano, variable = "tmean", period = "annual", trend = trn,
  title = "Annual Mean Temperature Anomaly — FWCP Peace Region"
)
```

```{r plot-prcp, fig.cap="Annual precipitation anomaly (% of 1951-1980 baseline) for the FWCP Peace Region."}
cd_plot_timeseries(
  ano, variable = "prcp", period = "annual", trend = trn,
  title = "Annual Precipitation Anomaly — FWCP Peace Region"
)
```

## Daytime Highs and Overnight Lows

The cd package ships daytime maximum (tmax) and overnight minimum (tmin)
temperatures alongside the daily mean. They carry distinct information.
Overnight minimums warming faster than daytime maximums — the
"day-night asymmetry" — is one of the textbook fingerprints of
greenhouse warming (Karl et al. 1993). Whether a watershed or region
shows that signal depends on local geography (valley inversions, snow
cover, slope-aspect mix).

```{r plot-tmax, fig.cap="Annual daytime maximum temperature (tmax) anomaly for the FWCP Peace Region relative to the 1951-1980 baseline."}
trn_tmax <- cd_trend(
  ano[ano$variable == "tmax" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "tmax", period = "annual", trend = trn_tmax,
  title = "Daytime Maximum (tmax) — Annual Anomaly")
```

```{r plot-tmin, fig.cap="Annual overnight minimum temperature (tmin) anomaly for the FWCP Peace Region relative to the 1951-1980 baseline."}
trn_tmin <- cd_trend(
  ano[ano$variable == "tmin" & ano$period == "annual", ],
  trend_start = c(1951, 1981)
)
cd_plot_timeseries(ano, variable = "tmin", period = "annual", trend = trn_tmin,
  title = "Overnight Minimum (tmin) — Annual Anomaly")
```

```{r plot-dtr, fig.cap="Diurnal temperature range (daytime maximum minus overnight minimum) annual mean for the FWCP Peace Region. The downward trend indicates overnight lows are warming faster than daytime highs — the textbook day-night asymmetry shows up here."}
tmax_ts <- ts[ts$variable == "tmax" & ts$period == "annual", c("year", "value")]
tmin_ts <- ts[ts$variable == "tmin" & ts$period == "annual", c("year", "value")]
dtr <- merge(tmax_ts, tmin_ts, by = "year", suffixes = c("_max", "_min"))
dtr$dtr <- dtr$value_max - dtr$value_min

ggplot(dtr, aes(x = year, y = dtr)) +
  geom_line(color = "grey50") +
  geom_point(color = "grey30", size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "#b2182b", linewidth = 0.8) +
  labs(
    title = "Diurnal Temperature Range — FWCP Peace Region",
    x = NULL,
    y = expression("Daytime maximum minus overnight minimum (" * degree * "C)")
  ) +
  theme_minimal(base_size = 12)
```

For the FWCP Peace Region, **overnight minimums are warming faster
than daytime maximums** — the textbook day-night asymmetry. Daytime
maximums warmed about +0.027 °C per year since 1951 (+2.0 °C
cumulative), while overnight minimums warmed about +0.032 °C per
year (+2.4 °C cumulative). The overnight side warmed roughly 0.4 °C
more than the daytime side over the full record, narrowing the
diurnal temperature range by the same amount.

## Recent vs Pre-warming

The tables below compare two windows directly — the recent decade
(2015–2025) against the pre-warming reference (1951–1980). Each row
includes the long-term Mann-Kendall p-value from the trend test (the
75-year window) so you can see whether the variable also has a
statistically significant monotonic trend. A small p-value (rule of
thumb: < 0.05) means the trend is unlikely to be noise. The trend test
asks a slightly different question than "are these two windows
different" — it tests for a steady directional change — but it is the
closest available proxy for now.

```{r compare-helpers, include = FALSE}
# Long-term Mann-Kendall p-values from the 1951 trend window, joined
# onto the compare tables so the reader can flag noisy variables.
trn_p <- trn[trn$trend_start == 1951, c("variable", "period", "mk_pvalue")]
names(trn_p)[3] <- "trend_p"
trn_p$trend_p <- round(trn_p$trend_p, 3)
```

```{r compare, tab.cap = "Recent (2015-2025) minus pre-warming (1951-1980) means. The trend_p column is the Mann-Kendall p-value for the 75-year (1951-present) trend on the same variable and period."}
# Equivalent to:
#   cmp <- cd_compare(ts, window_a = 2015:2025, window_b = 1951:1980,
#                     method = "mean_diff")
cmp <- merge(cmp, trn_p, by = c("variable", "period"), all.x = TRUE)
kableExtra::kable_styling(
  knitr::kable(cmp, caption = NA, digits = 2),
  bootstrap_options = c("striped", "hover", "condensed")
) |>
  kableExtra::scroll_box(height = "320px")
```

<br>

```{r compare-pct, tab.cap = "Percent change for precipitation and soil moisture, recent vs pre-warming. The trend_p column is the Mann-Kendall p-value for the 75-year trend."}
# Equivalent to:
#   cmp_pct <- cd_compare(
#     ts[ts$variable %in% c("prcp", "soil_moisture"), ],
#     window_a = 2015:2025, window_b = 1951:1980, method = "pct_change")
cmp_pct <- merge(cmp_pct, trn_p, by = c("variable", "period"), all.x = TRUE)
knitr::kable(cmp_pct, caption = NA, digits = 1)
```

<br>

## Spatial Pattern

The zonal mean reduces a region this size to a single number; the spatial
pattern carries the rest of the story.

```{r spatial-tmean, fig.cap="Spatial pattern of annual mean temperature departure across the FWCP Peace Region (2015-2025 mean minus 1951-1980 mean, degrees C).", fig.height=7}
# Pre-computed by data-raw/peace_fwcp_vignette_data.R. Live equivalent:
#   r_tmean <- cd_crop(catalog$href[catalog$variable == "tmean"
#                                   & catalog$period == "annual"], aoi)
#   years <- as.integer(names(r_tmean))
#   departure <- mean(r_tmean[[which(years >= 2015 & years <= 2025)]]) -
#                mean(r_tmean[[which(years >= 1951 & years <= 1980)]])
#   departure <- terra::mask(departure, aoi)
departure <- terra::rast(system.file(
  "vignette-data", "peace_fwcp_departure_tmean.tif", package = "cd"
))
names(departure) <- "Temperature departure"

ggplot() +
  geom_spatraster(data = departure) +
  geom_sf(data = ecoregions, fill = NA, color = "grey25",
          linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = aoi, fill = NA, color = "black", linewidth = 0.6) +
  geom_sf(data = lakes, fill = NA, color = "grey40", linewidth = 0.2) +
  geom_sf(data = highways, color = "#333333", linewidth = 0.4) +
  geom_sf(data = towns, color = "black", size = 2) +
  ggrepel::geom_label_repel(
    data = towns, aes(label = name, geometry = geom),
    stat = "sf_coordinates", size = 3, fill = "white", alpha = 0.8
  ) +
  scale_fill_distiller(
    palette = "RdBu", direction = -1,
    name = expression(Delta * degree * C)
  ) +
  coord_sf(
    xlim = c(aoi_bb["xmin"], aoi_bb["xmax"]),
    ylim = c(aoi_bb["ymin"], aoi_bb["ymax"])
  ) +
  labs(title = "Temperature Departure (2015-2025 vs 1951-1980)") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank())
```

## Per-Ecoregion Variation

The regional zonal mean averages over five ecoregions with different
elevations and exposures. To check whether the regional story holds
within each ecoregion — and where it does not — we run the same
pipeline on each ecoregion polygon individually.

```{r per-ecoregion-compute, include = FALSE}
# Pre-computed by data-raw/peace_fwcp_vignette_data.R. Live equivalent
# would loop cd_extract / cd_baseline / cd_anomaly / cd_trend / cd_compare
# over each ecoregion polygon.
results <- lapply(vignette_data$ecoregion, function(x) {
  list(ano = x$ano, trn = x$trn, cmp = x$cmp_pct)
})
# Reorder to match ecoregions (already by area_km2 desc on disk)
results <- results[ecoregions$code]

# Long-form labels: "OMM: Omineca Mountains"
er_labels <- setNames(
  paste0(ecoregions$code, ": ",
         tools::toTitleCase(tolower(ecoregions$name))),
  ecoregions$code
)

ano_all <- do.call(rbind, lapply(seq_along(results), function(i) {
  o <- results[[i]]$ano
  o$ecoregion <- ecoregions$code[i]
  o
}))
ano_all$ecoregion <- factor(ano_all$ecoregion, levels = ecoregions$code)

# Theil-Sen trend segments for facet overlay — one row per
# (ecoregion x trend_start) combination so both 75-yr and 45-yr lines
# render in each facet.
build_segments <- function(var) {
  do.call(rbind, lapply(seq_along(results), function(i) {
    sub <- results[[i]]$trn
    sub <- sub[sub$variable == var & sub$period == "annual", ]
    if (nrow(sub) == 0) return(NULL)
    yrs <- ano_all$year[ano_all$variable == var & ano_all$ecoregion == ecoregions$code[i]]
    data.frame(
      ecoregion = factor(ecoregions$code[i], levels = ecoregions$code),
      window = paste0(sub$trend_start, "–", max(yrs)),
      x_start = sub$trend_start,
      x_end = max(yrs),
      y_start = sub$intercept + sub$slope * sub$trend_start,
      y_end = sub$intercept + sub$slope * max(yrs)
    )
  }))
}
trn_segs_tmean <- build_segments("tmean")
trn_segs_prcp  <- build_segments("prcp")
```

```{r facet-tmean, fig.cap="Annual mean temperature anomaly relative to the 1951-1980 baseline, by ecoregion. Dashed line is the 75-year Theil-Sen trend (1951-present); solid line is the 45-year trend (1981-present). A solid line steeper than the dashed line indicates accelerating warming.", fig.height=6}
ggplot(ano_all[ano_all$variable == "tmean" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_tmean,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#d73027", `FALSE` = "#4575b4")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = expression("Mean temperature anomaly (" * degree * C * ")")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r facet-prcp, fig.cap="Annual precipitation anomaly (% of the 1951-1980 baseline) by ecoregion. Dashed line is the 75-year trend (1951-present); solid line is the 45-year trend (1981-present). The two northernmost ecoregions (BMP, NRM) show statistically significant precipitation increases over the full record; the others do not.", fig.height=6}
ggplot(ano_all[ano_all$variable == "prcp" & ano_all$period == "annual", ],
       aes(x = year, y = anomaly)) +
  geom_col(aes(fill = anomaly >= 0), width = 0.85, show.legend = FALSE) +
  geom_segment(data = trn_segs_prcp,
               aes(x = x_start, xend = x_end, y = y_start, yend = y_end,
                   linetype = window),
               inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(`TRUE` = "#4575b4", `FALSE` = "#d73027")) +
  scale_linetype_manual(values = c("dashed", "solid"), name = "Trend window") +
  facet_wrap(~ ecoregion, ncol = 3,
             labeller = labeller(ecoregion = er_labels)) +
  labs(x = NULL, y = "Precipitation anomaly (% of baseline)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r rollup, tab.cap = "Per-ecoregion roll-up over the 75-year window (1951-present): annual mean, daytime maximum, and overnight minimum temperature trends (degrees C per decade); annual precipitation trend (mm per year) and Mann-Kendall p-value; annual vapour pressure deficit trend (hPa per decade) and p-value; and recent (2015-2025) vs pre-warming (1951-1980) percent change for precipitation and soil moisture. All temperature p-values are below 0.001. A tmin slope greater than the tmax slope indicates the textbook day-night asymmetry."}
get_75 <- function(trn, var) {
  trn[trn$variable == var & trn$period == "annual" & trn$trend_start == 1951, ]
}
rollup <- do.call(rbind, lapply(seq_along(results), function(i) {
  res <- results[[i]]
  cmp <- res$cmp
  tmean <- get_75(res$trn, "tmean")
  tmax  <- get_75(res$trn, "tmax")
  tmin  <- get_75(res$trn, "tmin")
  prcp  <- get_75(res$trn, "prcp")
  vpd   <- get_75(res$trn, "vpd")
  data.frame(
    Ecoregion = ecoregions$code[i],
    `tmean degC/dec` = round(10 * tmean$slope, 2),
    `tmax degC/dec`  = round(10 * tmax$slope, 2),
    `tmin degC/dec`  = round(10 * tmin$slope, 2),
    `prcp mm/yr` = round(prcp$slope, 3),
    `prcp p` = round(prcp$mk_pvalue, 3),
    `vpd hPa/dec` = round(10 * vpd$slope, 3),
    `vpd p` = round(vpd$mk_pvalue, 3),
    `prcp pct change` = round(
      cmp$difference[cmp$variable == "prcp" & cmp$period == "annual"], 1
    ),
    `soil moisture pct change` = round(
      cmp$difference[cmp$variable == "soil_moisture" & cmp$period == "annual"], 1
    ),
    check.names = FALSE
  )
}))
knitr::kable(rollup, row.names = FALSE, caption = NA)
```

Ecoregion codes used in the table above: `r paste0(ecoregions$code, " — ", ecoregions$name_tc, collapse = "; ")`.

<br>

<br>

## Interpretation

Three findings emerge from the FWCP Peace Region trends.

**Warming is broad, fast, and uniform.** All five ecoregions warmed
between +1.7 °C and +1.9 °C cumulative since 1951, at a rate near
+0.3 °C per decade, with Mann-Kendall p-values below 0.001 in every
ecoregion. The trend is statistically significant beyond reasonable
doubt of being random noise. Daily maximum, daily minimum, and daily
mean temperatures all tell the same story. There is no thermal hot
spot. The whole region has moved together. A subtle gradient does
emerge — the northern, higher-elevation ecoregions (Omineca Mountains,
Boreal Mountains and Plateaus) warmed about 0.2 °C more than the
southern Fraser Basin, consistent with the well-documented pattern of
high-latitude and high-elevation amplification — but it is small
relative to the regional signal.

**Precipitation is increasing significantly only in the two
northernmost ecoregions.** Across the FWCP Peace as a whole, the
long-term trend in annual precipitation is not statistically
significant. Broken out by ecoregion, Boreal Mountains and Plateaus
and Northern Canadian Rocky Mountains — both in the northern half of
the region — show real increases in annual precipitation
(p = 0.02 and 0.01). The southern and central ecoregions show no
significant change. This is a north-versus-south contrast that the
regional average hides, and the place where breaking the analysis
down by ecoregion was worth doing.

**The atmosphere is drying despite, in places, more precipitation.**
Vapour pressure deficit — the gap between how much water the air
could hold and how much it actually does — rose significantly in
every ecoregion (p < 0.005 across all five). Warmer air holds more
water before saturating, and that pulls moisture out of soil and
vegetation through evaporation and transpiration. Soil moisture is
essentially flat in every ecoregion — even where precipitation
increased — because the warmer atmosphere is drinking the extra
water back. This is the headline ecological finding for the region:
water inputs may be rising in places, but soil and vegetation are
not seeing more available moisture.

For the cold-water salmonids the FWCP supports — bull trout, Arctic
grayling, mountain whitefish, rainbow trout — these signals point in
the same direction. Stream temperatures are likely rising in step with
warmer ambient air temperatures, and the evapotranspiration imbalance
means low-flow conditions in late summer are not being relieved by the
precipitation increase that did occur.
